---
title: "Systems Immunology Malaria Traveller Cohort"
subtitle: "Additional analysis"
author: "Maximilian Julius Lautenbach"
date: '`r format(Sys.Date(), "%Y-%m-%d")`'
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: kate
    code_folding: hide
    df_print: paged
knit: (function(inputFile, encoding) {
          rmarkdown::render(inputFile,
                            encoding = encoding, 
                            output_file = paste0(
                              xfun::sans_ext(inputFile), '_', Sys.Date(), '.html'),
                                output_dir = "../results/reports/")})
---

```{r setup, include=FALSE}
result.dir <- paste0("results/figures/")
options(stringsAsFactors = FALSE) 
```

## required libraries
```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(viridis)
library(RColorBrewer)
library(ComplexHeatmap)
library(circlize)
library(dagitty)
library(ggdag)
source("../scripts/helper_fun.R")
```

# Preparation

### load data

```{r message=FALSE, warning=FALSE}
all.data <- readRDS("../data/MEFISTO_input_data.Rds")
data.ab <- read_csv("../data/TravellerCohort_IgGclass_long.csv")
subjectTable <- read_csv("../data/TravellerCohort_SubjectTable.csv")
sampleTable <- read_csv("../data/TravellerCohort_SampleTable.csv")
```


## Figure S3_2 A
- T cell dynamics over time

```{r}
temp.data <- all.data %>% rename(sampleID = sample) %>% 
  inner_join(sampleTable,by="sampleID") %>% 
  filter(view=="Facs") %>%
  mutate(ID = gsub("\\|.*","",sampleID),
         Time = as.factor(gsub(".*\\|","",sampleID))) %>% 
  bind_rows(all.data %>% rename(sampleID = sample) %>% 
  inner_join(sampleTable,by="sampleID") %>% 
  filter(view=="Facs") %>%
  mutate(ID = gsub("\\|.*","",sampleID),
         Time = as.factor(gsub(".*\\|","",sampleID))) %>%  mutate(group="all_individuals")) %>% 
    mutate(group = factor(group, levels = c("all_individuals","primary_infected","previously_exposed"))) 
```


```{r message=FALSE, warning=FALSE, fig.width=3,fig.height=3}
activated_tcells <- c("CD8+CD38+HLADR+ T cells","CD8+CD57+CD38+ T cells","CD4+CD38+HLADR+ T cells","CD4+CD38+ T cells")
gd_tcells <- c("Total gdT cells","Vd2+ gdT cells","Vd2- gdT cells")

(p1 <- temp.data %>% 
  filter(feature %in% c(activated_tcells, gd_tcells)) %>% 

  ggplot(aes(x=weeks.po,y=value,color=group)) +
  geom_point(alpha=0.1,width = 0.25,size=0.8) +
  geom_line(aes(group=ID),size=0.5,alpha=0.125) +
  geom_smooth(aes(group=group),size=0.9,method = "loess") +
  scale_x_continuous(trans='log2') +
  facet_grid(feature~group,scales = "free",labeller = label_wrap_gen(width=10)) +
  scale_colour_manual(values = c(ENDEMIC_colors,"all_individuals"="red")) +
  labs(x="Weeks after symptom onset",
       y="cells/ul [log2]",
       caption = "loess fit with 0.95 confidence interval") +
  jtools::theme_nice() +
  theme_figpanel+
  theme(legend.position = "none",
        panel.spacing.x=unit(0.5, "lines"),
        panel.spacing.y=unit(0.5, "lines")))


#pdf("../plots/Figure_S3_2.pdf", width = 6,height = 7.5, useDingbats = F,   title="main", paper = "a4", compress = T ) #colormodel = "cmyk"
p1 + theme(legend.position = "none") 
#dev.off()#
```



## Figure S7 C-D
```{r}
temp.selection <- data.ab %>% select(sampleID,feature,value,Time,ID) %>% 
  bind_rows(all.data %>% rename(sampleID = sample) %>% 
              mutate(ID = gsub("\\|.*","",sampleID),
                     Time = as.factor(gsub(".*\\|","",sampleID))) %>% select(sampleID,feature,value,Time,ID)) %>%
  mutate(time_feature = paste0(Time,"_",feature)) %>% 
  select(-feature,-Time,-sampleID) %>% 
  spread(time_feature,value) 
```

```{r}
df <- temp.selection %>% inner_join(subjectTable %>% select(ID,Endemic,inf_rbc,cmv.status, hemoglobin_gt,years_since_endemic,years_in_endemic),by="ID") %>% mutate(hemoglobin_gt = factor(hemoglobin_gt, levels=c("AA","AS")),
                                                                                                                                cmv.status = factor(cmv.status, levels=c("negative","positive")),
                                                                                                                                Endemic = factor(Endemic, levels=c("primary_infected","previously_exposed")))
```

### S7C - IFNg

#### Minimal sufficient adjustment sets for estimating the total effect of *inf_RBC*on *ifng*:
* __igg3__
* __mono__

```{r}
dag <- dagitty('dag {
bb="0,0,1,1"
cmv [pos="0.633,0.144"]
exposure [pos="0.200,0.148"]
ifng [outcome,pos="0.198,0.504"]
igg3 [pos="0.502,0.501"]
inf_RBC [exposure,pos="0.420,0.690"]
memory [latent,pos="0.424,0.323"]
mono [pos="0.325,0.502"]
vd2gdt [pos="0.628,0.499"]
years_in_endemic [pos="0.340,0.147"]
years_since_endemic [pos="0.493,0.146"]
cmv -> memory
exposure -> memory
igg3 -> inf_RBC
inf_RBC -> ifng
memory -> ifng
memory -> igg3
memory -> mono
memory -> vd2gdt
mono -> inf_RBC
years_in_endemic -> memory
years_since_endemic -> memory
}')

dag_ifn_irbc <- ggdag_adjustment_set(dag, shadow = T, 
                     stylized = T,
                     node_size = 8,
                     text_size = 2,  
                     text_col = "black") + 
    theme_dag_grid(base_size = 6) +
  scale_color_manual(values = brewer.pal(5,"BrBG")[c(2,5)]) 
dag_ifn_irbc +  theme_dag_grid(base_size = 6) 
```

```{r}
# n=24
data4model <- df %>% select(`Acute_IFN-gamma`,Endemic,inf_rbc,Acute_IgG3,`Acute_Intermediate monocytes`) %>% na.omit()
m00 <- lm(`Acute_IFN-gamma` ~ Endemic, data = data4model)
m0 <- lm(`Acute_IFN-gamma` ~ inf_rbc, data = data4model)
m1 <- lm(`Acute_IFN-gamma` ~ inf_rbc + Acute_IgG3 + `Acute_Intermediate monocytes`, data = data4model)

modelsummary::modelsummary(list("PrevExposure" = m00,
                                "iRBC" = m0,
                                "iRBC | IgG3 + int. monocytes" = m1),
                           stars = T,
                            statistic = 'conf.int', 
             conf_level = .95)

modelsummary::modelplot(list("PrevExposure" = m00,
                             "iRBC" = m0,
                             "iRBC | IgG3 + int. monocytes" = m1),
                        stars =T, facet=F) +
  labs(x = 'Coefficients', 
       y = 'Term names',
       title = 'Linear regression models of IFNgamma (NPX)"',
       caption = "") +
  scale_color_manual(values = randomcoloR::distinctColorPalette(5))

performance::compare_performance(m00, m0,m1,rank = TRUE) 

# The Robust Likelihood Test (the LR column and its associated p-value) indicates whether each model fits better than the reference model.
anova(m00, m0,m1, test='LRT')

performance::test_likelihoodratio(m0,m1, estimator = "OLS")
```

### S7D - D10 Vd2+ gdTcells

### Minimal sufficient adjustment sets for estimating the total effect of inf_RBC on vd2gdt:
* __igg3__
* __mono__

```{r}
dag <- dagitty('dag {
bb="0,0,1,1"
cmv [pos="0.633,0.144"]
exposure [pos="0.200,0.148"]
ifng [pos="0.198,0.504"]
igg3 [pos="0.502,0.501"]
inf_RBC [exposure,pos="0.420,0.690"]
memory [latent,pos="0.424,0.323"]
mono [pos="0.325,0.502"]
vd2gdt [outcome,pos="0.628,0.499"]
years_in_endemic [pos="0.340,0.147"]
years_since_endemic [pos="0.493,0.146"]
cmv -> memory
exposure -> memory
igg3 -> inf_RBC
inf_RBC -> vd2gdt
memory -> ifng
memory -> igg3
memory -> mono
memory -> vd2gdt
mono -> inf_RBC
years_in_endemic -> memory
years_since_endemic -> memory
}')

impliedConditionalIndependencies(dag)

dag_gdt_irbc <- ggdag_adjustment_set(dag, shadow = T, 
                     stylized = T,
                     node_size = 8,
                     text_size = 2, 
                     text_col = "black") + 
  theme_dag_grid()+
  scale_color_manual(values = brewer.pal(5,"BrBG")[c(2,5)])

dag_gdt_irbc + theme_dag_grid(base_size = 6) 
```
```{r}
# n=9
data4model <- df %>% select(`D10_Vd2+ gdT cells`,Endemic,inf_rbc,Acute_IgG3,Acute_IgG1,`Acute_Intermediate monocytes`,cmv.status,hemoglobin_gt) %>% na.omit()
m00 <- lm(`D10_Vd2+ gdT cells` ~ Endemic, data = data4model)
m0 <- lm(`D10_Vd2+ gdT cells` ~ inf_rbc, data = data4model)
m1 <- lm(`D10_Vd2+ gdT cells` ~ inf_rbc + Acute_IgG3 + `Acute_Intermediate monocytes`, data = data4model)
m2 <- lm(`D10_Vd2+ gdT cells` ~ inf_rbc + Acute_IgG1 + `Acute_Intermediate monocytes`, data = data4model)


modelsummary::modelsummary(list("PrevExposure" = m00,
                                "iRBC" = m0,
                                "iRBC | IgG3 + int. monocytes" = m1),
                           stars = T, statistic = 'conf.int', 
             conf_level = .95)

modelsummary::modelsummary(list("PrevExposure" = m00,
                                "iRBC" = m0,
                                "iRBC | IgG1 + int. monocytes" = m2),
                           stars = T, statistic = 'conf.int', 
             conf_level = .95)

modelsummary::modelplot(list("PrevExposure" = m00,
                             "iRBC" = m0,
                             "iRBC | IgG3 + int. monocytes" = m1),
                        stars =T, facet=F) +
  labs(x = 'Coefficients', 
       y = 'Term names',
       title = 'Linear regression models of Vd2+ gdT cells"',
       caption = "") +
  scale_color_manual(values = randomcoloR::distinctColorPalette(3))

performance::compare_performance(m00, m0,m1,m2, rank = TRUE) 

# The Robust Likelihood Test (the LR column and its associated p-value) indicates whether each model fits better than the reference model.
anova(m00, m0,m1, test='LRT')
anova(m00, m0,m2, test='LRT')

performance::test_likelihoodratio(m0,m1, estimator = "OLS")
performance::test_likelihoodratio(m0,m2, estimator = "OLS")
```

## save DAG for S7 C, D
```{r}
library(patchwork)
#pdf("../plots/Figure_S7c_d.pdf", width = 6, height = 8, useDingbats = T, title="Supplementary7", bg="transparent",# paper = "a4",colormodel = "cmyk", compress = T, pagecentre =T)

(dag_ifn_irbc + theme(legend.position = "none")) /
    (dag_gdt_irbc + theme(legend.position = "none"))

#dev.off()
```

## Figure S7 E-H
### Does parasitemia impact immune response 

```{r message=FALSE, warning=FALSE}
set.seed(2022)
k3 <- kmeans(subjectTable %>% select(ID,inf_rbc) %>% column_to_rownames("ID") %>% mutate(inf_rbc=log(inf_rbc)), centers=3)# %>% mutate(k3 = kmeans(inf_rbc, centers = 3))

cluster.lab <- as.data.frame(k3$centers) %>% rownames_to_column("cluster") %>% arrange(inf_rbc) %>% mutate(cluster_lab = c("low","medium","high"))

inf_rbc_cluster <- 
  data.frame(cluster = k3$cluster) %>% rownames_to_column("ID") %>% 
  inner_join(subjectTable %>% select(ID,inf_rbc,Endemic)) %>% 
  right_join(cluster.lab %>% select(cluster,cluster_lab) %>% mutate(cluster = as.numeric(cluster)), by ="cluster") %>% 
  mutate(cluster = factor(cluster_lab, levels=c("low","medium","high"))) 
```


```{r message=FALSE, warning=FALSE}
selected.group = "medium"

inf_rbc_cluster %>% filter(cluster == selected.group) %>% count(Endemic)

selection_irbc_IDs <- inf_rbc_cluster %>% filter(cluster == selected.group) %>% pull(ID)

## get Ab data
selection_irbc_ab.data <- data.ab %>% filter(ID %in% selection_irbc_IDs)

## get protein/cell data
temp <-  all.data %>% rename(sampleID = sample) %>% 
  inner_join(sampleTable,by="sampleID") %>% 
  mutate(ID = gsub("\\|.*","",sampleID),
         Time = as.factor(gsub(".*\\|","",sampleID)))

selection_irbc_protein.data <- temp %>% filter(view=="Olink", ID %in% selection_irbc_IDs)
selection_irbc_cell.data <- temp %>% filter(view=="Facs", ID %in% selection_irbc_IDs)

temp <- bind_rows(
  selection_irbc_ab.data %>% select(sampleID,feature,value,Time,ID),
  selection_irbc_protein.data %>% select(sampleID,feature,value,Time,ID)) %>% 
  bind_rows(selection_irbc_cell.data %>% select(sampleID,feature,value,Time,ID))

temp.selection <- temp %>% mutate(time_feature = paste0(Time,"_",feature)) %>% 
  filter(time_feature %in% c("Acute_IFN-gamma","Acute_TNF","Acute_IgG3","Acute_Intermediate monocytes","D10_Vd2+ gdT cells")) %>% 
  select(-feature,-Time,-sampleID) %>% 
  mutate(time_feature = gsub("_"," ",time_feature)) %>% 
  spread(time_feature,value) %>% 
  inner_join(inf_rbc_cluster,by="ID") 

##########################

(p1 <- inf_rbc_cluster %>% 
   ggplot(aes(x= cluster,y=inf_rbc, fill=Endemic)) +
   geom_jitter(width = 0.25,size=1,shape=21, alpha=1) +
   scale_fill_manual(values = ENDEMIC_colors) +
    ylim(0,20) +
   labs(x="",
        y="iRBC (%)",
        fill="",
        ) +
    theme12 +
  jtools::theme_nice() +
  theme_figpanel+
    theme(legend.position = "none"))

######################

(p2 <- temp.selection %>%  
   gather(feature,value,-c("Endemic","cluster","ID")) %>% 
   mutate(value=as.double(value)) %>% 
   filter(feature %in% c("Acute IFN-gamma","Acute TNF")) %>% 
   ggviolin(x = "Endemic",
            y = "value",
            color = "Endemic", 
            width = 0.6,
            size=0.2,
            add = c("jitter","boxplot"),
            add.params = list(width=0.4, alpha = 0.5,size=0.2)) + 
   theme(legend.position = "none") +
   facet_grid(.~feature,scales = "free",labeller = label_wrap_gen(width=10)) +
   scale_color_manual(values=ENDEMIC_colors) +
   stat_compare_means(label.y = 19,label.x = 1.3, method = "wilcox.test",label = "p.format",size = 2,
                      paired = F)  +   # Add global p-value 
   ylim(0,21) +
   labs(x="",
        y="NPX",
        #caption = "Mann-Whitney test"
   ) +
     jtools::theme_nice() +
  theme_figpanel+
   theme(legend.position = "none",
     axis.text.x = element_blank(),
     axis.ticks.x = element_blank(),
     panel.spacing.x=unit(0.5, "lines"),
        panel.spacing.y=unit(0.5, "lines")))

######################

(p3 <- temp.selection %>%  
   gather(feature,value,-c("Endemic","cluster","ID")) %>% 
      mutate(value=as.double(value)) %>% 
   filter(feature %in% c("Acute Intermediate monocytes","D10 Vd2+ gdT cells")) %>%
   ggviolin(x = "Endemic",
            y = "value",
            color = "Endemic", 
            width = 0.6,
            size=0.2,
            add = c("jitter","boxplot"),
            add.params = list(width=0.4, alpha = 0.5,size=0.2)) + 
   facet_grid(.~feature,scales = "free",labeller = label_wrap_gen(width=10)) +
   scale_color_manual(values=ENDEMIC_colors) +
   stat_compare_means(label.y = 11,label.x = 1.3, method = "wilcox.test",label = "p.format",size = 2,
                      paired = F)  +   # Add global p-value 
   ylim(0,12) +
   labs(x="",
        y="cells/ul (log2)",
        #caption = "Mann-Whitney test"
   ) +
jtools::theme_nice() +
  theme_figpanel+
   theme(legend.position = "none",
     axis.text.x = element_blank(),
     axis.ticks.x = element_blank(),
     panel.spacing.x=unit(0.5, "lines"),
        panel.spacing.y=unit(0.5, "lines")))

######################

(p4 <- temp.selection %>%  
   gather(feature,value,-c("Endemic","cluster","ID")) %>% 
      mutate(value=as.double(value)) %>% 

   filter(feature %in% c("Acute IgG3")) %>% 
   ggviolin(x = "Endemic",
            y = "value",
            color = "Endemic", 
            size=0.2,
            width = 0.6,
            add = c("jitter","boxplot"),
            add.params = c(list(width=0.2, alpha = 0.5,size=0.2))) + 
   facet_grid(.~feature,scales = "free") +
   scale_color_manual(values=ENDEMIC_colors) +
   stat_compare_means(label.y = 1,label.x = 1.3, method = "wilcox.test",label = "p.format",size = 2,
                      paired = F)  +   # Add global p-value 
   ylim(0,1.1) +
   labs(x="",
        y="IgG subclass CRS",
        #caption = "Mann-Whitney test"
   ) +
   theme_classic(base_size = 6) +
    jtools::theme_nice() +
  theme_figpanel+
   theme(legend.position = "none",
         axis.text.x = element_blank(),
         axis.ticks.x = element_blank()))
```


```{r message=FALSE, warning=FALSE}
library(patchwork)
#pdf("../plots/Figure_S7e_h.pdf", width = 6,height = 2, useDingbats = F,   title="main", paper = "a4", compress = T ) 
p1 + theme(legend.position = "none") + 
  p2 + theme(#strip.text.x = element_text(size = 3.5),
    axis.text.x = element_blank(),
                              axis.ticks.x = element_blank()) +
  p3 + theme(
    axis.text.x = element_blank(),
                              axis.ticks.x = element_blank()) +
  p4 +  theme(
    axis.text.x = element_blank(),
                              axis.ticks.x = element_blank()) +
  plot_layout(widths = c(1.25,2,2,1))
#dev.off()#
```






