---
title: "Systems Immunology Malaria Traveller Cohort"
subtitle: "Downstream correlation analysis"
author: "Maximilian Julius Lautenbach"
date: '`r format(Sys.Date(), "%Y-%m-%d")`'
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: kate
    code_folding: hide
    df_print: paged
knit: (function(inputFile, encoding) {
          rmarkdown::render(inputFile,
                            encoding = encoding, 
                            output_file = paste0(
                              xfun::sans_ext(inputFile), '_', Sys.Date(), '.html'),
                                output_dir = "../")})
---

```{r setup, include=FALSE}
knitr::opts_knit$set(echo = TRUE, root.dir = getwd(), fig.width = 6, fig.height = 5, warning = FALSE, message = FALSE)
options(stringsAsFactors = FALSE) 
```

## required libraries
```{r message=FALSE}
library(tidyverse)
library(correlation)
library(circlize)
library(ComplexHeatmap) # needed for legends
library(igraph)
library(ggraph)
source("helper_fun.R")
```

# Preparation

## Importing data
* selected features from MEFISTO model, generated with ```MEFISTO_downstream.Rmd```
* IgG subclass Cummulative Response Score data, generated with ```CRScore.Rmd```

```{r message=FALSE, warning=FALSE}
subjectTable <- read_csv("../data/TravellerCohort_SubjectTable.csv")
selected.features.proteins <- read_csv("../data/Top10_Proteins_F1.csv")
selected.features.cells <- read_csv("../data/Top10_Cells_F1.csv") 
mefisto.input <- readRDS("../data/MEFISTO_input_data.Rds")
data.ab <- read_csv("../data/TravellerCohort_IgGclass_long.csv")
```

# Correlation analysis

## Figure 3

### Correlations data prep

```{r message=FALSE, warning=FALSE}
corr.input <- mefisto.input %>% 
  filter(feature %in% c(read_csv("../data/Top10_Cells_F1.csv") %>% pull(feature),
                        read_csv("../data/Top10_Proteins_F1.csv")  %>% pull(feature))) %>% 
  mutate(Time = gsub(".*\\|","",sample),
         feature = paste0(view,"|",Time,"|",feature),
         ID = gsub("\\|.*","",sample)) %>% 
  select(ID,feature,value) %>% 
  spread(feature,value)

corr.output <- correlation::correlation(data = corr.input,
                                        method = "spearman",
                                        p_adjust = "fdr",
                                        ci = 0.95,
                                        redundant = T)

```

* filter for  acute time point cytokines and post acute immune cell subsets

```{r}
corr.output.filtered <- corr.output %>% 
  as.tibble() %>% 
  filter(grepl('Olink',Parameter1), grepl("Acute",Parameter1)) %>% 
  filter(grepl("Facs",Parameter2), grepl("D10|M1|M3|M6|Y1",Parameter2)) %>% 
  arrange(desc(abs(rho)),p)
corr.output.filtered
```

### Fig. 3A 
* Overall correlation of pcute proteins with post-acute immune cell subsets
* no __rho__ or __p value__ threshold 

```{r fig.asp=1, message=FALSE, warning=FALSE}
input <- corr.output.filtered %>% 
  select(Parameter1,Parameter2,rho) %>% 
  mutate(Parameter2 = gsub("Facs\\|","",Parameter2)) %>% 
  spread(Parameter2,rho) %>% 
  replace(is.na(.), 0) %>% 
  mutate(Parameter1 = gsub("Olink\\|Acute\\|","",Parameter1)) %>% 
  column_to_rownames("Parameter1") %>% 
  t()  %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  mutate(Time = as.factor(gsub("\\|.*","",rowname))) %>% 
  select(rowname, Time, everything()) %>% 
  arrange(Time) 

## set row/column annotation
row.anno <- data.frame(rowname = input$rowname,
                       SampleTimePoint = input$Time) %>%
  column_to_rownames("rowname")

col.anno <- data.frame(colname = names(input)[3:length(names(input))],
                       SampleTimePoint = "Acute") %>%
  column_to_rownames("colname")

## plot heatmap
(hm <- input %>% 
    column_to_rownames("rowname") %>% 
    select(-Time) %>% 
    ComplexHeatmap::pheatmap(cluster_rows = T, 
                             cluster_cols = T,
                             show_colnames = T, 
                             show_rownames = T,
                             annotation_names_row = F,
                             name = "Spearman rho",
                             annotation_colors = list("SampleTimePoint" = TIME_colors),  
                             annotation_legend = T,
                             scale = "none",
                             labels_row = gsub(".*\\|","",input$rowname),
                             annotation_row = row.anno,
                             annotation_col = col.anno,
                             cellwidth = 10, 
                             cellheight = 9,
                             color = colorRamp2(c(-1,0,1), c("#4575b4","#ffffbf","#d73027"))))
#pdf(file=paste0("../plots/Factor1_AcuteProt2PostAcuteCells_Cor_heatmap_all.pdf"), width=11, height = 15)
hm
#dev.off()
```


### Fig. 3B 
* Corddiagram of significant correlations
* threshold __rho__ 0.7
* threshold __p value__ 0.05

#### Color and legend settings
```{r cord_plot_prep}
col_fun = colorRamp2(c(-1,0,1), c("#4575b4","#ffffbf","#d73027"))
lgd_rho <- Legend(col_fun = col_fun, title = "Spearman rho \n",
                  legend_height = unit(0.9, "snpc"),
                  grid_width = unit(0.2, "snpc"),
                  labels_gp = gpar(fontsize = 30),
                  title_gp = gpar(fontsize = 30)) ## dependend on ComplexHeatmap
```

#### Corddiagram
```{r message=FALSE, warning=FALSE, fig.asp=1}
df <- corr.output %>% 
  as.tibble() %>% 
  filter(grepl("Olink",Parameter1), grepl("Acute",Parameter1)) %>% 
  filter(grepl("Facs",Parameter2), grepl("D10|M1|M3|M6|Y1",Parameter2)) %>% 
  arrange(desc(abs(rho)),p) %>% 
  filter(p <= 0.05, abs(rho) > 0.7) %>% 
  select(Parameter1,Parameter2,rho) %>% 
  rename(from = Parameter1, to = Parameter2, value = rho) %>% 
  mutate(from = gsub("Olink\\|","",from),
         to = gsub("Facs\\|","",to)) %>% 
  rename(r = value)

sections <- as.data.frame(union(df[[1]], df[[2]]))
colnames(sections) <- "union" 
sections <- sections %>%
  mutate(Time = gsub("\\|.*","",union),
         feature = gsub(".*\\|","",union)) %>%
  arrange(Time, feature)
##============================================================# 
##          CordDiagram                                       #
##============================================================#
#pdf(file=paste0("../plots/Factor1Top20_AcuteProt_PostCells_significant_corddiagram.pdf"), width=30, height = 30, pointsize = 10)

chordDiagram(df, 
             annotationTrack = "",
             order = sections$union,
             big.gap = 15,
             small.gap = 2,
             col=col_fun(df$r), 
             scale = TRUE,
             preAllocateTracks = list(
               ## 1 (outer track for space)
               list(bg.border = NA,#"black",
                    bg.col=NULL,
                    track.height = 0.4),
               #2 (outer)
               list(ylim = c(0, 1),
                    bg.col = NA,
                    bg.border = NA,
                    track.height = 0.0001,
                    bg.lty = par("lty"),
                    bg.lwd = par("lwd"),
                    big.gap = 10),
               #3 (inner)
               list(bg.border = NA,#"black",
                    bg.col=NULL,
                    track.height = 0.1)))

circos.trackPlotRegion(track.index = 2, 
                       bg.border = NA,
                       panel.fun = function(x, y) {
                         xlim = get.cell.meta.data("xlim")
                         ylim = get.cell.meta.data("ylim")
                         sector.name = sub(".*\\|","", get.cell.meta.data("sector.index"))
                         
                         circos.text(mean(xlim), ylim[1] + 1.1,
                                     sector.name, 
                                     facing = "clockwise", 
                                     niceFacing = TRUE,
                                     cex = 3,
                                     adj = c(-0.05, 0.5), 
                                     col = "black")
                       })

highlight.sector(c(sections[sections$Time == "Acute",]$union), 
                 track.index = 3,
                 col = "#C51B7D",
                 text = "Acute",
                 cex = 3,
                 text.col = "black",
                 niceFacing = TRUE)
highlight.sector(c(sections[sections$Time == "D10",]$union), 
                 track.index = 3,
                 col =  "#E9A3C9",
                 text = "D10",
                 cex = 3,
                 text.col = "black",
                 niceFacing = TRUE)
highlight.sector(c(sections[sections$Time == "M3",]$union), 
                 track.index = 3,
                 col = "#E6F5D0",
                 text = "M3",
                 cex = 3,
                 text.col = "black",
                 niceFacing = TRUE)
highlight.sector(c(sections[sections$Time == "M6",]$union), 
                 track.index = 3,
                 col =  "#A1D76A",
                 text = "M6",
                 cex = 3,
                 text.col = "black",
                 niceFacing = TRUE)
highlight.sector(c(sections[sections$Time == "Y1",]$union), 
                 track.index = 3,
                 col =  "#4D9221" ,
                 text = "Y1",
                 cex = 3,
                 text.col = "black",
                 niceFacing = TRUE)

text(x = unit(0.8, "snpc"), # snpc unit gives you a square region,
     y = unit(-0.97, "snpc"),
     paste0("\n","FDR < 0.05","\n","| rho | > 0.7"), cex = 3)
draw(lgd_rho, 
     x = unit(1, "snpc"), # snpc unit gives you a square region,
     y = unit(0.25, "snpc"),
     just = "right")

#dev.off()
circos.clear()
```

### Fig. 3C-G

#### Scatterplot

```{r message=FALSE, warning=FALSE}
corr.output.sigA_D10 <- corr.output %>% 
  as.tibble() %>% 
  filter(grepl("Olink",Parameter1), grepl("Acute",Parameter1)) %>% 
  filter(grepl("Facs",Parameter2), grepl("D10",Parameter2)) %>% 
  filter(Parameter1 != Parameter2) %>% 
  arrange(desc(abs(rho)),p) %>% 
  filter(p <= 0.05)

scatter.fig.folder <- "/CorrelationScatter/"
ifelse(isFALSE(dir.exists(paste0("../plots",scatter.fig.folder))),
       dir.create(paste0("../plots",scatter.fig.folder), recursive = TRUE),
       paste0(scatter.fig.folder," Folder excists exists already!"))

for(position in c(1:length(corr.output.sigA_D10$Parameter1))){
  selection <- corr.output.sigA_D10 %>% 
    slice(position) %>% 
    select(Parameter1,Parameter2) %>% unlist() %>% unname()
  
  statistic_annotation <- corr.output.sigA_D10 %>% slice(position)
  
  plot.input <- corr.input %>% select(ID,selection) %>% 
    na.omit() %>% inner_join(subjectTable %>% select(ID,Endemic),by="ID") %>% column_to_rownames("ID")
  
  plot <- plot.input %>% 
    ggplot(aes(x = .[,1], y = .[,2])) +
    coord_flip() +
    geom_point(col="black",size=3, alpha = 0.7) +
    labs(x = paste("NPX \n",selection[1]),
         y = paste0(selection[2],"\n cells per uL blood (log2)")) +
    geom_smooth(colour = "black", method = "lm", se = FALSE, alpha = 0.7) +
    scale_color_manual(values=ENDEMIC_colors) +
    labs(subtitle = paste0(" rho = ", round(statistic_annotation$rho, 2), 
                           ", p.adj = ",round(statistic_annotation$p, 3),
                           #", CI ",round(statistic_annotation$CI_low,2),"-",round(statistic_annotation$CI_high,2),
                           ", n obs. = ",round(statistic_annotation$n_Obs, 3)))  +
    theme_classic() +
    theme12 +
    theme(legend.position='none',
          plot.subtitle=element_text(size=10, hjust=0.5, face="bold", color="black"))
  #ggsave(paste0("../plots/",scatter.fig.folder,"CorrScatter_Rank",position,"_",selection[2],"_",selection[1],".pdf"),plot, width = 3.5, height = 3.5)
}
```

## Figure 5

### Fig. 5B 
* Corddiagram of significant gdTcells ~acute protein associations only
* threshold __rho__ 0.7
* threshold __p value__ 0.05 

```{r message=FALSE, warning=FALSE}
df <- corr.output %>% 
  as.tibble() %>% 
  filter(grepl("Olink",Parameter1), grepl("Acute",Parameter1)) %>% 
  filter(grepl("Facs",Parameter2), grepl("D10|M1|M3|M6|Y1",Parameter2)) %>% 
  filter(grepl("gdT cell", Parameter2)) %>% 
  arrange(desc(abs(rho)),p) %>% 
  filter(p <= 0.05) %>% 
  select(Parameter1,Parameter2,rho) %>% 
  rename(from = Parameter1, to = Parameter2, value = rho) %>% 
  mutate(from = gsub("Olink\\|","",from),
         to = gsub("Facs\\|","",to)) %>% 
  rename(r = value)

sections <- as.data.frame(union(df[[1]], df[[2]]))
colnames(sections) <- "union" 
sections <- sections %>%
  mutate(Time = gsub("\\|.*","",union),
         feature = gsub(".*\\|","",union)) %>%
  arrange(Time, feature)
##============================================================# 
##          CordDiagram                                       #
##============================================================#
#pdf(file=paste0("../plots/gdTcells_AcuteProt_PostCells_significant_corddiagram.pdf"), width=30, height = 30, pointsize = 10)
chordDiagram(df, 
             annotationTrack = "",#c("grid"),
             order = sections$union,
             big.gap = 50,
             small.gap = 2,
             ## connection color
             col=col_fun(df$r), 
             scale = TRUE,
             preAllocateTracks = list(
               ## 1 (outer track for space)
               list(bg.border = NA,#"black",
                    bg.col=NULL,
                    track.height = 0.4),
               #2 (outer)
               list(ylim = c(0, 1),
                    bg.col = NA,
                    bg.border = NA,
                    track.height = 0.0001,
                    bg.lty = par("lty"),
                    bg.lwd = par("lwd"),
                    big.gap = 10),
               
               #3 (inner)
               list(bg.border = NA,#"black",
                    bg.col=NULL,
                    track.height = 0.1)))

circos.trackPlotRegion(track.index = 2, 
                       bg.border = NA,
                       panel.fun = function(x, y) {
                         xlim = get.cell.meta.data("xlim")
                         ylim = get.cell.meta.data("ylim")
                         sector.name = sub(".*\\|","", get.cell.meta.data("sector.index"))
                         
                         circos.text(mean(xlim), ylim[1] + 1.1,
                                     sector.name, 
                                     facing = "clockwise", 
                                     niceFacing = TRUE,
                                     cex = 3,
                                     adj = c(-0.05, 0.5), 
                                     col = "black")
                       })

highlight.sector(c(sections[sections$Time == "Acute",]$union), 
                 track.index = 3,
                 col = "#C51B7D",
                 text = "Acute",
                 cex = 3,
                 text.col = "black",
                 niceFacing = TRUE)
highlight.sector(c(sections[sections$Time == "D10",]$union), 
                 track.index = 3,
                 col =  "#E9A3C9",
                 text = "D10",
                 cex = 3,
                 text.col = "black",
                 niceFacing = TRUE)

text(x = unit(0.8, "snpc"), # snpc unit gives you a square region,
     y = unit(-0.97, "snpc"),
     paste0("\n","FDR < 0.05"), cex = 3)
draw(lgd_rho, 
     x = unit(1, "snpc"), # snpc unit gives you a square region,
     y = unit(0.25, "snpc"),
     just = "right",
)
#dev.off()
circos.clear()

```
# Figure 7

## Correlation of IgG subclass data

```{r message=FALSE, warning=FALSE}
corr.ab.input <- corr.input %>% 
  full_join(data.ab %>% 
              mutate(feature = paste0("CRS","|",Time,"|",feature)) %>% 
              select(ID,feature,value) %>% 
              spread(feature,value),
            by="ID") %>% select(ID,contains("Acute"),contains("D10")) %>% 
  column_to_rownames("ID")
```

```{r}
corr.ab.output <- correlation::correlation(data = corr.ab.input,
                                           method = "spearman",
                                           p_adjust = "fdr",
                                           ci = 0.95,
                                           redundant = T)
```

## Acut Protein ~ Acute CRS

### Fig. 7B
* threshold __p value__ 0.05 

```{r}
selection.sig <- corr.ab.output %>% 
  as.tibble() %>% 
  filter(grepl("CRS",Parameter1), grepl("Acute",Parameter1)) %>% 
  filter(grepl("Facs|Olink",Parameter2), grepl("Acute",Parameter2)) %>% 
    filter(Parameter1 != Parameter2) %>% 

  arrange(desc(abs(rho)),p) %>%
  #filter(abs(rho)>0.7) %>% 
  filter(p <= 0.05)
```



```{r fig.asp=1}
#mat <- corr.ab.output %>% 
#  as.tibble() %>% 
#  filter(grepl("Olink",Parameter1), grepl("Acute",Parameter1)) %>% 
#  filter(grepl("CRS",Parameter2), grepl("Acute",Parameter2))  %>% #grepl("D10|M1|M3|M6|Y1",Parameter2)) %>% 
#  filter(Parameter1 != Parameter2) %>% 
#  arrange(desc(abs(rho)),p) %>% 
#  filter(p <= 0.05) %>% 
#  select(Parameter1,Parameter2,rho,p)
 
mat <- selection.sig %>% 
  select(Parameter1,Parameter2,rho,p)

## define graph input
vert <-  mat %>%
  rename(x=Parameter1,
         y=Parameter2,
         r=rho) %>% 
  select(x,y,r,p) %>% gather(item, node,x:y) %>%
  group_by(node) %>% count(node) %>% 
  mutate(to.split = node) %>% 
  separate(to.split, sep="\\|", into=c("View","Time","feature"))

graph_cors <- mat %>%
  rename(x=Parameter1,
         y=Parameter2,
         r=rho) %>% 
  select(x,y,r,p) %>% 
  graph_from_data_frame(directed = FALSE, vert)

## option 2 via layout
layout <- create_layout(graph_cors, layout = 'igraph', algorithm = 'kk') 



(gg <- 
    ggraph(layout) +
  geom_edge_link(aes(edge_alpha = abs(r), 
                     color = r),
  edge_width = 2) +
  guides(edge_alpha = "none", edge_width = "none") +
  geom_node_point(aes(color = Time,
                      size = n),
                  show.legend = T) +
  guides(size = "none") +
  labs(edge_color="Spearman rho") +
  
  scale_color_manual(values = TIME_colors) +
  scale_edge_color_gradientn(limits = c(-1, 1), colors = c("#4575b4","#ffffbf","#d73027")) +
  geom_node_text(aes(label = feature),
                 size=10,
                 repel = TRUE) +
  theme_graph() +
  scale_size(range = c(4,10)) +
  labs(title = "Correlations of acute IgG subclass response and pro-inflammatory signature",
       caption = "FDR < 0.05"))

#ggsave(paste0("../plots/Acute_IgG_Subclass_CytokineCorrelation_graph.pdf"),gg, width = 10, height = 10)

```


### Fig. 7C
#### Scatterplot

```{r}
## create folder
scatter.fig.folder <- "/CorrelationScatter_wIgG_Acute/"
ifelse(isFALSE(dir.exists(paste0("../plots",scatter.fig.folder))),
       dir.create(paste0("../plots",scatter.fig.folder), recursive = TRUE),
       paste0(scatter.fig.folder," Folder excists exists already!"))

## loop to create scatterplots for significant correlations
for(position in c(1:length(selection.sig$Parameter1))){
  selection <- selection.sig %>% 
    slice(position) %>% 
    select(Parameter1,Parameter2) %>% unlist() %>% unname()
  
  statistic_annotation <- selection.sig %>% slice(position)
  
  plot.input <- corr.ab.input %>% select(selection) %>% 
    na.omit() %>% rownames_to_column("ID") %>% inner_join(subjectTable %>% select(ID,Endemic),by="ID") %>% column_to_rownames("ID")
  
  plot <- plot.input %>% 
    ggplot(aes(x = .[,1], y = .[,2])) +
    coord_flip() +
    geom_point(aes(col=Endemic),
               size=3, alpha = 0.7) +
    labs(x = paste(selection[1]),
         y = paste0(selection[2])) +
    geom_smooth(colour = "black", method = "lm", se = FALSE, alpha = 0.7) +
    scale_color_manual(values=ENDEMIC_colors) +
    labs(subtitle = paste0(" rho = ", round(statistic_annotation$rho, 2), 
                           ", p.adj = ",round(statistic_annotation$p, 3),
                           ", n obs. = ",round(statistic_annotation$n_Obs, 3)))  +
    theme_classic() +
    theme12 +
    theme(legend.position='none',
          plot.subtitle=element_text(size=10, hjust=0.5, face="bold", color="black"))
 # ggsave(paste0("../plots",scatter.fig.folder,"CorrScatter_Rank",position,"_",selection[2],"_",selection[1],".pdf"),plot, width = 3.5, height = 3.5)
}

```



```{r}
corr.ab.output.sig <- corr.ab.output %>% 
  as.tibble() %>% 
  filter(grepl("CRS",Parameter1), grepl("Acute",Parameter1)) %>% 
  filter(grepl("Facs",Parameter2), grepl("D10",Parameter2)) %>% 
  arrange(desc(abs(rho)),p) %>%
  filter(abs(rho)>0.7,
         p <= 0.05) 
```

## Acut CRS/Protein ~ D10 cells
 
### Fig. 7D
* threshold __rho__ 0.7
* threshold __p value__ 0.05 

```{r}
#mat <- corr.ab.output %>% 
#  as.tibble() %>% 
#  filter(grepl("CRS",Parameter1), grepl("Acute",Parameter1)) %>% 
#  filter(grepl("Facs",Parameter2), grepl("D10",Parameter2))  %>% #grepl("D10|M1|M3|M6|Y1",Parameter2)) %>% 
#  filter(Parameter1 != Parameter2) %>% 
#  arrange(desc(abs(rho)),p) %>% 
#  filter(p <= 0.05,
#         abs(rho) > 0.7) %>% 
  
mat <- corr.ab.output.sig %>% 
  select(Parameter1,Parameter2,rho,p)

## define graph input
vert <-  mat %>%
  rename(x=Parameter1,
         y=Parameter2,
         r=rho) %>% 
  select(x,y,r,p) %>% gather(item, node,x:y) %>%
  group_by(node) %>% count(node) %>% 
  mutate(to.split = node) %>% 
  separate(to.split, sep="\\|", into=c("View","Time","feature"))

graph_cors <- mat %>%
  rename(x=Parameter1,
         y=Parameter2,
         r=rho) %>% 
  select(x,y,r,p) %>% 
  graph_from_data_frame(directed = FALSE, vert)


## option 2 via layout
layout <- create_layout(graph_cors, layout = 'igraph', algorithm = 'kk') #, algorithm = 'kk' #layout = "centrality",cent = graph.strength(graph_cors))#

gg <- ggraph(layout) +
  geom_edge_link(aes(edge_alpha = abs(r), 
                     #edge_width = 1-p,
                     color = r
  ),
  edge_width = 2) +
  guides(edge_alpha = "none", edge_width = "none") +
  #geom_edge_link0(aes(color = r, 
  #                    edge_width = 1-p),
  #                show.legend = F) +
  geom_node_point(aes(color = Time,
                      size = n),
                  show.legend = T) +
  guides(size = "none") +
  scale_color_manual(values = TIME_colors) +
  scale_edge_color_gradientn(limits = c(-1, 1), colors = c("#4575b4","#ffffbf","#d73027")) +
  geom_node_text(aes(label = feature),
                 size=10,
                 repel = TRUE) +
  theme_graph() +
  scale_size(range = c(4,10)) +

  labs(title = "Correlations of Acute IgG subclass response and D10 cells",
       caption = "FDR < 0.05",
       edge_color="Spearman rho")
# ggsave(paste0("../plots/Acute_IgG_Subclass_D10cells_graph_fdr005.pdf"),gg, width = 10, height = 10)

```

### Fig. 7E
#### Scatterplot

```{r message=FALSE, warning=FALSE}
## create folder
scatter.fig.folder <- "/CorrelationScatter_wIgG/"
ifelse(isFALSE(dir.exists(paste0("../plots",scatter.fig.folder))),
       dir.create(paste0("../plots",scatter.fig.folder), recursive = TRUE),
       paste0(scatter.fig.folder," Folder excists exists already!"))

## loop to create scatterplots for significant correlations
for(position in c(1:length(corr.ab.output.sig$Parameter1))){
  selection <- corr.ab.output.sig %>% 
    slice(position) %>% 
    select(Parameter1,Parameter2) %>% unlist() %>% unname()
  
  statistic_annotation <- corr.ab.output.sig %>% slice(position)
  
  plot.input <- corr.ab.input %>% select(selection) %>% 
    na.omit() %>% rownames_to_column("ID") %>% inner_join(subjectTable %>% select(ID,Endemic),by="ID") %>% column_to_rownames("ID")
  
  plot <- plot.input %>% 
    ggplot(aes(x = .[,1], y = .[,2])) +
    coord_flip() +
    geom_point(aes(col=Endemic),
               size=3, alpha = 0.7) + 
    labs(x = paste(selection[1]),
         y = paste0(selection[2])) +
    geom_smooth(colour = "black", method = "lm", se = FALSE, alpha = 0.7) +
    scale_color_manual(values=ENDEMIC_colors) +
    labs(subtitle = paste0(" rho = ", round(statistic_annotation$rho, 2), #statistic_annotation$Method, 
                           ", p.adj = ",round(statistic_annotation$p, 3),
                           #", CI ",round(statistic_annotation$CI_low,2),"-",round(statistic_annotation$CI_high,2),
                           ", n obs. = ",round(statistic_annotation$n_Obs, 3)))  +
    theme_classic() +
    theme12 +
    theme(legend.position='none',
          plot.subtitle=element_text(size=10, hjust=0.5, face="bold", color="black"))
  
#  ggsave(paste0("../plots",scatter.fig.folder,"CorrScatter_Rank",position,"_",selection[2],"_",selection[1],".pdf"),plot, width = 3.5, height = 3.5)
  }

```




## =================
## How does IgG4 correlates in previously exposed group?

```{r message=FALSE, warning=FALSE}
corr.ab.input.group <- corr.input %>% 
  full_join(data.ab %>% 
              mutate(feature = paste0("CRS","|",Time,"|",feature)) %>% 
              select(ID,feature,value) %>% 
              spread(feature,value),
            by="ID") %>% 
  select(ID,contains("Acute"),contains("D10")) %>% 
  inner_join(subjectTable %>% select(ID,Endemic), by="ID") %>% 
 filter(Endemic =="previously_exposed") %>% 
  select(-Endemic) %>% 
  column_to_rownames("ID")

corr.ab.output.group <- correlation::correlation(data = corr.ab.input.group,
                                           method = "spearman",
                                           p_adjust = "fdr",
                                           ci = 0.95,
                                           redundant = F)


corr.ab.output.group.sig <- corr.ab.output.group %>% 
  as.tibble() %>% 
  filter(grepl("CRS|Olink",Parameter1), grepl("Acute",Parameter1)) %>% 
  #filter(grepl("Facs",Parameter2), grepl("D10",Parameter2)) %>% 
  arrange(desc(abs(rho)),p) %>%
  filter(abs(rho)>0.7) %>% 
  filter(p <= 0.05) 
```



```{r}
sessionInfo()
```


