---
title: "Systems Immunology Malaria Traveller Cohort"
subtitle: "Data re-purposing of study by Yman et al. 2019"
author: "Maximilian Julius Lautenbach"
date: '`r format(Sys.Date(), "%Y-%m-%d")`'
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: kate
    code_folding: hide
    df_print: paged
knit: (function(inputFile, encoding) {
          rmarkdown::render(inputFile,
                            encoding = encoding, 
                            output_file = paste0(
                              xfun::sans_ext(inputFile), '_', Sys.Date(), '.html'),
                                output_dir = "../results/reports/")})
---

```{r setup, include=FALSE}
result.dir <- paste0("../results/figures/")
options(stringsAsFactors = FALSE) 
```


## required libraries
```{r message=FALSE}
library(tidyverse)
library(readxl)
source("helper_fun.R")
```

# Preparation

### load data 
```{r message=FALSE, warning=FALSE}
ab.data.raw <- read_csv("../data/YmanEtAl2019_adj_lev_raw_data.csv")
subjectTable <- read_csv("../data/TravellerCohort_SubjectTable.csv")
sampleTable <- read_csv("../data/TravellerCohort_SampleTable.csv")
ID.list <- read_csv("../../Mefisto/data/PatientIDs_includedinMofaModel.csv") %>% rename(subjectID = "x") %>% pull(subjectID)
```

### rank normalise data

```{r}
ab.norm <- ab.data.raw %>% 
  group_by(igg,antigen) %>% 
  mutate(rnorm.value = rank_norm_fun(adj_lev_raw),
         zscore = scale(rnorm.value)) %>% 
  ungroup()
```

### wrangel data

```{r}
ab.norm <- ab.norm %>% 
  mutate(Time = as.factor(gsub(".*\\|","",sampleID)),
         subjectID = gsub("\\|.*","",sampleID)) %>% 
  rename(IgG_class = igg,
         antigen_full = antigen) %>% 
  mutate(
    ## unique antigens
    antigen = ifelse(grepl("AMA1", antigen_full),"AMA1",
                     ifelse(grepl("MSP1", antigen_full),"MSP1",
                            ifelse(grepl("MSP2", antigen_full),"MSP2",
                                   ifelse(grepl("MSP3", antigen_full),"MSP3",
                                          ifelse(grepl("RH5", antigen_full),"RH5",
                                                 ifelse(grepl("Schizont", antigen_full),"Schizont",
                                                        ifelse(grepl("TTd", antigen_full),"TTd",antigen_full))))))),
    cat = ifelse(antigen == "TTd","ctrl.antigen","test.antigen"),
    include_in_ic_sum = ifelse(antigen == "TTd",F,
                               ifelse(antigen == "Schizont",F,T)))
```


### calculate Cummulative Response Score (CRS)
* sum of rank-normalized values per Individual, Time point and IgG class
* range normalization for values between 0 and 1

```{r}
ab.norm.out <- ab.norm %>% 
  filter(cat == "test.antigen") %>% ## filter out control antigen TTd 
  filter(include_in_ic_sum == T) %>% 
  group_by(subjectID,Time, IgG_class) %>% 
  summarise(n = n(),
            ic.sum = sum(zscore)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  group_by(IgG_class) %>% 
  mutate(ic.sum = range01(ic.sum)) %>% 
  ungroup() %>% 
  mutate(sampleID = paste0(subjectID,"|",Time)) %>% 
  rename(value = ic.sum,
         feature = IgG_class) %>% 
  dplyr::select(sampleID,feature,value) %>% 
  mutate(Omics = "Antibody",
         unite = "IgG class [sum of zscores]") %>% 
  mutate(Time = as.factor(gsub(".*\\|","",sampleID)),
         ID = gsub("\\|.*","",sampleID)) 

## save data for further analysis
#write.csv(ab.norm.out, paste0("../data/TravellerCohort_IgGclass_long.csv"), row.names = FALSE)
```
```{r fig.asp=1}
(qc.plot <- ab.norm %>% 
   filter(cat == "test.antigen",
          antigen != "Schizont") %>% 
   ggplot(aes(x=rnorm.value, fill= antigen_full)) +
   geom_density(alpha=.3) +
   facet_grid(antigen ~ IgG_class,cols = ,scales = "free_y") +
   labs(x="rank normalized value",
        fill = "") +
   geom_vline(xintercept=0,
              linetype="dashed",
              size=1, colour="red") +
   theme(legend.position = "top") +
   jtools::theme_nice() +
   theme_figpanel +
   theme(legend.position = "top",
         panel.spacing.y=unit(0.5, "lines"),
         panel.spacing.x=unit(0.5, "lines")))
```



## Figure 7

### Fig. 7A
* Plot CRS over time
* requires function **longitudinal_timepoint_plot()** from ```helper_fun.R```

```{r fig.asp=0.2}
source("../scripts/helper_fun.R")
p1 <- ab.norm.out %>% 
    inner_join(subjectTable %>% select(ID,Endemic), by="ID") %>% 
    filter(ID %in% ID.list) %>% 
    rename(group = Endemic) %>% 
    longitudinal_timepoint_plot() +
    facet_wrap(~feature, ncol = 5) + 
    labs(y="Scaled IgG subclass \n response score",
         subtitle = "Cumulative response to multiple blood stage antigens") +
    scale_y_continuous(limits=c(0,1)) +
    jtools::theme_nice() +
    theme_figpanel +
    theme(legend.position = "none",
          plot.subtitle = element_text(size=6),
          axis.title.y = element_text(vjust = -3),
          panel.spacing.x=unit(0.5, "lines"))
#pdf("../plots/CRS_overtime_revision_feb15.pdf", width = 5.25,height = 1.5, useDingbats = F,   title="main", paper = "a4", compress = T ) 
p1 
#dev.off()#

```


### Fig. S7A
### qc plot 
* normal distribution required for subsequent cumulative score calculation


```{r fig.asp=1}
#pdf("../plots/IgGsubclassRespAntigens_QC_revision_feb7.pdf", width = 4,height = 4, useDingbats = F,   title="main", paper = "a4", compress = T ) 
qc.plot 
#dev.off()
```
### Fig. S7B 

```{r fig.asp=0.5}
(p1 <- ab.norm.out %>% 
   inner_join(subjectTable %>% select(ID,Endemic), by="ID") %>% 
   filter(ID %in% ID.list) %>% 
   ggplot(aes(x = Time, y = value, color=Endemic, fill = Endemic , group = interaction(Time,Endemic)))+
   geom_boxplot(position = position_dodge(1), alpha = 6/10, color="grey20",outlier.size = 0.5,
                width=0.5) +
   geom_dotplot(binaxis='y', 
                stackdir='center',
                dotsize = 0.2,
                alpha = 3/10,
                show.legend = F,
                position = position_dodge(1)) +
   scale_fill_manual(values=ENDEMIC_colors) +
   scale_color_manual(values=ENDEMIC_colors) +
   jtools::theme_nice() +
   facet_wrap(~feature,ncol = 5) +
   labs(x="Sample Time Points",
        y="Scaled IgG subclass \n response score") +
   jtools::theme_nice() +
   theme_figpanel +
   theme(panel.spacing.x=unit(0.5, "lines"),
         legend.position = "bottom",
         axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5)))

#pdf("../plots/IgGsubclassRespAntigens_s5_revision_feb7.pdf", width = 6,height = 3, useDingbats = F,   title="main", paper = "a4", compress = T ) 
p1 
#dev.off()
```

