---
title: "Systems Immunology Malaria Traveller Cohort"
subtitle: "Internal & External Comparison"
author: "Maximilian Julius Lautenbach"
date: '`r format(Sys.Date(), "%Y-%m-%d")`'
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: kate
    code_folding: hide
    df_print: paged
knit: (function(inputFile, encoding) {
          rmarkdown::render(inputFile,
                            encoding = encoding, 
                            output_file = paste0(
                              xfun::sans_ext(inputFile), '_', Sys.Date(), '.html'),
                                output_dir = "../results/reports/")})
---

```{r setup, include=FALSE}
result.dir <- paste0("../results/figures/")
options(stringsAsFactors = FALSE) 
```

## required libraries
```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(viridis)
library(ggpubr)
library(rstatix)
library(gridExtra)
library(cowplot)
library(RColorBrewer)
source("helper_fun.R")
```

## load data

```{r message=FALSE, warning=FALSE}
mefisto.input <- readRDS("../data/MEFISTO_input_data.Rds")
all.data <- readRDS("../data/ALL_input_data.Rds")

facs.pkl.data <- read_csv("../data/TravellerCohort_FACS_log2pKlive_long.csv")
facs.cpu.data <- read_csv("../data/TravellerCohort_FACS_log2cpu_long.csv")
subjectTable <- read_csv("../data/TravellerCohort_SubjectTable.csv")
sampleTable <- read_csv("../data/TravellerCohort_SampleTable.csv")

uniprot.keywords <- read.csv("../data/UniprotKeywordMapping.csv")
wbc.data <- read_csv("../data/TravellerCohort_WBC_long.csv")
protein.data.ctrl <- read_csv("../results/2021-08-23/2021-08-23_Olink_HC_data_QC.csv")
facs.pkl.data.ctrl <- read_csv("../results/2021-08-23/2021-08-23_FACS_data_HC_selection.csv")

mefisto.selection.proteins <- read_csv("../data/Top10_Proteins_F1.csv")
mefisto.selection.cells <- read_csv("../data/Top10_Cells_F1.csv")
```


# == Internal comparison - Acute vs Convalescence == 

```{r fig.show="hold", message=FALSE, warning=FALSE, out.width="50%"}
input_temp <- all.data %>% 
  mutate(Time = as.factor(gsub(".*\\|","",sample)),
         ID = gsub("\\|.*","",sample)) 

data_acute_conv <- input_temp %>% 
  select(ID,view,Time,value,feature) %>% 
  filter(Time == "Acute") %>% 
  spread(Time,value) %>% 
  full_join(input_temp %>% 
              filter(Time %in% c("M6","Y1")) %>% 
              group_by(ID, feature) %>%
              summarise(Convalescence = mean(value)) %>% 
              ungroup(), 
            by=c("ID","feature")) %>% 
  gather(Acute_Conv, value, -ID,-feature,-view) %>% 
  inner_join(subjectTable,by="ID") %>%  
  select(ID,Endemic, Acute_Conv, view,feature, value) %>% 
  mutate(unite = ifelse(view == "Olink","NPX","cells per uL blood (log2)"),
         Acute_Conv = factor(Acute_Conv, levels = c("Acute", "Convalescence")),
         Endemic = factor(Endemic, levels = c("primary_infected","previously_exposed")))
head(data_acute_conv)
```

## Figure Supplementary 2

### Fig. S2D-E
* Acute vs Convalscence top10
* Normalise to a convalesecence median for both groups together

#### Positive Factor1 values

```{r message=FALSE, warning=FALSE}
factor1_ordered_prot_pos <- mefisto.selection.proteins %>% 
  arrange(desc(value)) %>% 
  select(feature,value) %>%
  head(10) %>% 
  rename(factor1.value = value)

medianConv <- data_acute_conv %>% 
  filter(Acute_Conv == "Convalescence") %>% 
  group_by(feature) %>% ## if Endemic add a few lines below too (right join)
  summarise(ConvValue = median(value, na.rm = T)) %>% 
  ungroup()

ac_norm.data <- right_join(data_acute_conv, medianConv, by=c("feature")) %>% #"Endemic",
  mutate(norm2ConvValue = value - ConvValue) %>% 
  mutate(Acute_Conv = factor(Acute_Conv, levels = c("Acute","Convalescence"))) 

## == Olink
plot.data <- ac_norm.data %>% 
  filter(view == "Olink") %>%
  dplyr::select(ID,Endemic,Acute_Conv,view,feature,norm2ConvValue,unite)

plot.data.ordered <-  factor1_ordered_prot_pos %>% 
  inner_join(plot.data, by="feature")

## ======== Acute vs Convalescence ====== ##
stat.test <- plot.data.ordered %>%
  group_by(feature) %>%
  wilcox_test(norm2ConvValue ~ Acute_Conv, p.adjust.method = "fdr") %>%
  adjust_pvalue() %>%
  add_significance("p.adj") %>% 
  inner_join(plot.data.ordered %>% select(feature,factor1.value) %>% distinct(), by="feature") %>% 
  arrange(desc(factor1.value)) %>% 
  add_xy_position(x="factor1.value")

conv.norm <-
  plot.data.ordered %>% 
  ggplot(aes(x= fct_reorder(feature, factor1.value), y= norm2ConvValue)) + 
  geom_hline(yintercept= 0, color = "grey20") +
  geom_boxplot(aes(fill=Acute_Conv),
               width = 1, 
               position = position_dodge(1),
               outlier.size = 0.2) +
  scale_fill_manual(values = c("#ca5572","grey50")) +
  labs(y = paste0(plot.data.ordered$unite," (normalized to convalescence median)"),
       x = "",
       title = "",
       caption = "Wilcoxon Tests with FDR correction") +
  theme_classic() +
  coord_flip() +
  theme_pubr() +
  theme(legend.position = "bottom") +
  stat_pvalue_manual(stat.test,
                     label = "p.adj.signif", 
                     y.position = min(plot.data.ordered$norm2ConvValue, na.rm = T)*1.1,
                     remove.bracket = T) 
(acute1.prot <- conv.norm)
#conv.norm %>% ggsave(file = paste0(result.dir,"Olink_Factor1_positive_Top10_Conv_normalized.pdf"),  width = 6, height = 7, useDingbats = FALSE)

## == Facs
factor1_ordered_cells_pos <- mefisto.selection.cells %>% 
  arrange(desc(value)) %>% 
  select(feature,value) %>%
  head(10) %>% 
  rename(factor1.value = value) 

plot.data <- ac_norm.data %>% 
  filter(view == "Facs") %>%
  dplyr::select(ID,Endemic,Acute_Conv,view,feature,norm2ConvValue,unite)


plot.data.ordered <- factor1_ordered_cells_pos %>% 
  inner_join(plot.data, by="feature")

## ======== Acute vs Convalescence ====== ##
stat.test <- plot.data.ordered %>%
  group_by(feature) %>%
  wilcox_test(norm2ConvValue ~ Acute_Conv, p.adjust.method = "fdr") %>%
  adjust_pvalue() %>%
  add_significance("p.adj") %>% 
  inner_join(plot.data.ordered %>% select(feature,factor1.value) %>% distinct(), by="feature") %>% 
  arrange(desc(factor1.value)) %>% 
  add_xy_position(x="factor1.value")

conv.norm <-
  plot.data.ordered %>% 
  ggplot(aes(x= fct_reorder(feature, factor1.value), y= norm2ConvValue)) + 
  geom_hline(yintercept= 0, color = "grey20") +
  geom_boxplot(aes(fill=Acute_Conv),
               width = 1, 
               position = position_dodge(1),
               outlier.size = 0.2) +
  scale_fill_manual(values = c("#ca5572","grey50")) +
  labs(y = paste0(plot.data.ordered$unite," (normalized to convalescence median)"),
       x = "",
       title = "",
       caption = "Wilcoxon Tests with FDR correction") +
  theme_classic() +
  coord_flip() +
  theme_pubr() +
  theme(legend.position = "bottom") +
  stat_pvalue_manual(stat.test,
                     label = "p.adj.signif", 
                     y.position = min(plot.data.ordered$norm2ConvValue, na.rm = T)*1.1,
                     remove.bracket = T) 
(acute1.cell <- conv.norm)
#conv.norm %>% ggsave(file = paste0(result.dir,"Facs_Factor1_positive_Top10_Conv_normalized.pdf"),  width = 8, height = 7, useDingbats = FALSE)

```

### Fig. S2F-G
 Acute vs Convalscence top10
* Normalise to a convalesecence median for both groups together

#### Negative Factor1 values

```{r message=TRUE, warning=FALSE}
factor1_ordered_prot_neg <- mefisto.selection.proteins %>% 
  arrange(value) %>% 
  select(feature,value) %>%
  head(10) %>% 
  rename(factor1.value = value)

medianConv <- data_acute_conv %>% 
  filter(Acute_Conv == "Acute") %>% 
  group_by(feature) %>% 
  summarise(ConvValue = median(value, na.rm = T)) %>% 
  ungroup()

ac_norm.data <-  right_join(data_acute_conv, medianConv, by=c("feature")) %>% #"Endemic",
  mutate(norm2ConvValue = value - ConvValue) %>% 
  mutate(Acute_Conv = factor(Acute_Conv, levels = c("Acute","Convalescence"))) 

## == Olink
plot.data <- ac_norm.data %>% 
  filter(view == "Olink") %>%
  dplyr::select(ID,Endemic,Acute_Conv,view,feature,norm2ConvValue,unite)

plot.data.ordered <-  factor1_ordered_prot_neg %>% 
  inner_join(plot.data, by="feature")

## ======== Acute vs Convalescence ====== ##
stat.test <- plot.data.ordered %>%
  group_by(feature) %>%
  wilcox_test(norm2ConvValue ~ Acute_Conv, p.adjust.method = "fdr") %>%
  adjust_pvalue() %>%
  add_significance("p.adj") %>% 
  inner_join(plot.data.ordered %>% select(feature,factor1.value) %>% distinct(), by="feature") %>% 
  arrange(desc(abs(factor1.value))) %>% 
  add_xy_position(x="factor1.value")

conv.norm <-
  plot.data.ordered %>% 
  ggplot(aes(x= fct_reorder(feature, factor1.value), y= norm2ConvValue)) + 
  geom_hline(yintercept= 0, color = "grey20") +
  geom_boxplot(aes(fill=Acute_Conv),
               width = 1, 
               position = position_dodge(1),
               outlier.size = 0.2) +
  scale_fill_manual(values = c("#ca5572","grey50")) +
  labs(y = paste0(plot.data.ordered$unite," (normalized to acute median)"),
       x = "",
       title = "",
       caption = "Wilcoxon Tests with FDR correction") +
  theme_classic() +
  coord_flip() +
  theme_pubr() +
  theme(legend.position = "bottom") +
  stat_pvalue_manual(stat.test,
                     label = "p.adj.signif", 
                     y.position = min(plot.data.ordered$norm2ConvValue, na.rm = T)*1.1,
                     remove.bracket = T) 

(conv1.prot <- conv.norm)
#conv.norm %>% ggsave(file = paste0(result.dir,"Olink_Factor1_negative_Top10_Acute_normalized.pdf"),  width = 6, height = 7, useDingbats = FALSE)


## Negative Factor1 values cells
factor1_ordered_cells_neg <- mefisto.selection.cells %>% 
  arrange(value) %>% 
  select(feature,value) %>%
  head(10) %>% 
  rename(factor1.value = value) 

plot.data <- ac_norm.data %>% 
  filter(view == "Facs") %>%
  dplyr::select(ID,Endemic,Acute_Conv,view,feature,norm2ConvValue,unite)


plot.data.ordered <- factor1_ordered_cells_neg %>% 
  inner_join(plot.data, by="feature") 

## ======== Acute vs Convalescence ====== ##
stat.test <- plot.data.ordered %>%
  group_by(feature) %>%
  wilcox_test(norm2ConvValue ~ Acute_Conv, p.adjust.method = "fdr") %>%
  adjust_pvalue() %>%
  add_significance("p.adj") %>% 
  inner_join(plot.data.ordered %>% select(feature,factor1.value) %>% distinct(), by="feature") %>% 
  arrange(desc(abs(factor1.value))) %>% 
  add_xy_position(x="factor1.value")

conv.norm <-
  plot.data.ordered %>% 
  ggplot(aes(x= fct_reorder(feature, factor1.value), y= norm2ConvValue)) + 
  geom_hline(yintercept= 0, color = "grey20") +
  geom_boxplot(aes(fill=Acute_Conv),
               width = 1, 
               position = position_dodge(1),
               outlier.size = 0.2) +
  scale_fill_manual(values = c("#ca5572","grey50")) +
  labs(y = paste0(plot.data.ordered$unite," (normalized to acute median)"),
       x = "",
       title = "",
       caption = "Wilcoxon Tests with FDR correction") +
  theme_classic() +
  coord_flip() +
  theme_pubr() +
  theme(legend.position = "bottom") +
  stat_pvalue_manual(stat.test,
                     label = "p.adj.signif", 
                     y.position = min(plot.data.ordered$norm2ConvValue, na.rm = T)*1.1,
                     remove.bracket = T) 
(conv1.cell <- conv.norm)
#conv.norm %>% ggsave(file = paste0(result.dir,"Facs_Factor1_negative_Top10_Acute_normalized.pdf"),  width = 8, height = 7, useDingbats = FALSE)
```

#### Uniprot kewords associated with proteins
* Keywords assigned to proteins because they are involved in a particular biological process.

```{r}
acute.prot2 <- uniprot.keywords %>% 
  right_join(factor1_ordered_prot_pos %>% 
               rename(Protein = feature) , by="Protein") %>% 
  filter(!keyword %in% c("Angiogenesis","Notch signaling pathway","Innate immunity")) %>% 
  spread(keyword,status) %>% 
  replace(is.na(.), 0) %>% 
  #select(-'<NA>') %>% 
  gather(keyword, status,-Protein,-factor1.value) %>%  
  ggplot(aes(x = keyword, y = reorder(Protein,factor1.value), fill = as.factor(status))) +
  geom_tile(color = "#ffffff") +
  scale_fill_manual(values= c("lightgrey","#1B9E77")) +
  theme_classic() +
  coord_equal() +
  theme_pubr() +
  theme(axis.title = element_blank(),
        legend.position = "none",
        axis.text.x.top = element_text(angle = 45, hjust=0)) +
  scale_x_discrete(position = "top") +
  scale_y_discrete(position = "right") 

acute.prot2 #%>% ggsave(file = paste0(result.dir,"Factor1_positive_Olink_UniProtKeywords_alternative.pdf"), width = 6, height = 7, useDingbats = FALSE)


conv.prot2 <- uniprot.keywords %>% 
  right_join(factor1_ordered_prot_neg %>% 
               rename(Protein = feature) , by="Protein") %>% #distinct(Protein)
  mutate(status = ifelse(is.na(keyword),0,status)) %>% 
  spread(keyword,status) %>% 
  replace(is.na(.), 0) %>% 
  select(-'<NA>') %>% 
  gather(keyword, status,-Protein,-factor1.value) %>%  
  filter(!keyword %in% c(NA,"Notch signaling pathway","Innate immunity","Immunity")) %>% 
  ggplot(aes(x = keyword, y = reorder(Protein,factor1.value), fill = as.factor(status))) +
  geom_tile(color = "#ffffff") +
  scale_fill_manual(values= c("lightgrey","#1B9E77")) +
  theme_classic() +
  coord_equal() +
  theme_pubr() +
  theme(axis.title = element_blank(),
        legend.position = "none",
        axis.text.x.top = element_text(angle = 45, hjust=0)) +
  scale_x_discrete(position = "top") +
  scale_y_discrete(position = "right") 
conv.prot2# %>% ggsave(file = paste0(result.dir,"Factor1_negative_Olink_UniProtKeywords_alternative.pdf"), width = 6, height = 7, useDingbats = FALSE)
```

#### scoring cells based on their markers

```{r message=FALSE, warning=FALSE}
cell.activation <- mefisto.selection.cells %>% 
  select(feature) %>% 
  mutate(Activation = ifelse(grepl("CD38\\+HLADR\\+",feature),"yes",
                             ifelse(grepl("CD38\\+",feature),"yes","no")),
         Proliferation = ifelse(grepl("CD38\\+HLADR\\+",feature),"yes","no"),
         Differentiation = ifelse(grepl("Non-classical", feature),"yes",
                                  ifelse(grepl("Temra", feature),"yes","no")),
         Exhaustion = ifelse(grepl("CD56-CD16+", feature),"yes",
                             ifelse(grepl("CD8+CD57+", feature),"yes",
                                    ifelse(grepl("CD8+CD57+ Temra",feature), "yes", "no")))) %>% 
  gather(status.lab, status,-feature)

cell.activation.plot <- function(arranged_data){
  arranged_data %>% 
    inner_join(cell.activation, by="feature") %>% 
    ggplot(aes(x = status.lab, y = reorder(feature, factor1.value), fill = as.factor(status))) +
    geom_tile(color = "#ffffff") +
    scale_fill_manual(values= c("lightgrey","#7570B3" )) +
    theme_classic() +
    coord_equal() +
    theme_pubr() +
    theme(axis.title = element_blank(),
          legend.position = "none",
          axis.text.x.top = element_text(angle = 45, hjust=0)) +
    scale_x_discrete(position = "top") +
    scale_y_discrete(position = "right")
}

acute.cell2 <- factor1_ordered_cells_pos %>% 
  cell.activation.plot()  
acute.cell2 #%>% ggsave(file = paste0(result.dir,"Factor1_positive_Facs_CellstatusKeywords_alternative.pdf"), width = 6, height = 7, useDingbats = FALSE)


conv.cell2 <- factor1_ordered_cells_neg %>% 
  cell.activation.plot() 

conv.cell2 #%>% ggsave(file = paste0(result.dir,"Factor1_negative_Facs_CellstatusKeywords_alternative.pdf"), width = 6, height = 7, useDingbats = FALSE)
```


```{r}
library(patchwork)
theme_complex <- theme(axis.text.y = element_blank(),
                       axis.ticks.y = element_blank(), 
                       axis.title.x = element_blank(),
                       axis.title.y = element_blank())
change.xy <- theme(axis.text.x = element_text(size = 5),
                  axis.text.y = element_text(size = 5))


#pdf("../plots/InternalComp_panel.pdf", width = 7,height = 3.5,useDingbats = F, title="Supplementary", paper = "a4", compress = T, ) #colormodel = "cmyk"
#par(mfrow = c(2,2))

acute1.cell + theme_figpanel + change.xy | acute.cell2 + theme_figpanel + change.xy + theme_complex  | acute1.prot + theme_figpanel + change.xy | acute.prot2 + theme_figpanel + change.xy + theme_complex 

conv1.cell + theme_figpanel + change.xy | conv.cell2 + theme_figpanel + change.xy + theme_complex | conv1.prot + theme_figpanel+ change.xy  | conv.prot2 + theme_figpanel + change.xy + theme_complex

#dev.off()
```


# == External comparison - Cohort vs Healthy Controls ==

## Supplementary Figure S3
* Immune cell abundance and plasma protein abundance compared to healthy controls
* Immune cell abundance based on cells/1000 live cell counts
* for each sample time point


```{r}
HC_protein.data <- protein.data.ctrl %>% 
  mutate(ID = gsub("_","",ID),
         view = Omics,
         group = "Ctrl",
         Acute_Conv = "Ctrl",
         unite = "NPX") %>% 
  dplyr::select(ID,group,Acute_Conv,view,feature,value,unite)

## Nonendemic_Endemic_Control data 
NEC <- bind_rows(HC_protein.data, data_acute_conv) %>%
  filter(view == "Olink") %>% 
  mutate(group = as.factor(group),
         Acute_Conv = as.factor(Acute_Conv)) %>% 
  mutate(group = factor(group, levels = c("primary_infected","previously_exposed","Ctrl")))

```


##### Data prep Immune cells

```{r}
## cohort data

Endemic_cell_data <- facs.pkl.data %>% 
  rename(view = Omics) %>% 
  mutate(Time = as.factor(gsub(".*\\|","",sampleID)),
         ID = gsub("\\|.*","",sampleID))  %>% 
  select(ID,view,Time,value,feature) %>% 
  inner_join(subjectTable, by="ID") %>% 
  select(ID,Endemic, Time, view, feature, value) %>% 
  mutate(#unite = ifelse(view == "Olink","NPX","cells per uL blood (log2)"),
         Endemic = factor(Endemic, levels = c("primary_infected","previously_exposed"))) %>% 
  mutate_if(is.character,
            stri_trans_general,
            id = "latin-ascii")

## calculate median of healthy ctrl
HC_cell.data <- facs.pkl.data.ctrl %>% 
  filter(gate != "id") %>% 
  mutate(ID = gsub("_","",sampleID),
         feature = gate.lab,
         view = "Facs",
         Endemic = "Ctrl",
         Time = "Ctrl",
         unite = "pkl") %>% 
  dplyr::select(ID,Endemic,Time,view,feature,value,unite) %>% 
  mutate(log.value = log(value),
         log2.value = log2(value +1),
         log10.value = log10(value)) %>% 
  select(ID,Endemic,Time,view,feature,log2.value) %>% 
  rename(value=log2.value) %>% 
  mutate_if(is.character,
            stri_trans_general,
            id = "latin-ascii")
  

NEC.cell <- rbind(Endemic_cell_data, HC_cell.data) %>%
  filter(view == "Facs") %>% 
  mutate(Endemic = as.factor(Endemic),
         Time = as.factor(Time)) %>% 
  mutate(Endemic = factor(Endemic, levels = c("primary_infected","previously_exposed","Ctrl")))

## join ctrl medians with patient data
cell.data.hc.normalized <- NEC.cell %>% 
  full_join(NEC.cell %>% 
              filter(Time == "Ctrl") %>%
              group_by(feature) %>% 
              summarise(hcpkl = median(value, na.rm = T)), # healthy control per k (1000) live cells
            by = "feature") %>% 
  mutate(HealthyInf = as.factor(ifelse(Endemic == "Ctrl","Healthy","Malaria infected"))) %>% 
  mutate(norm2hcpkl = value - hcpkl) %>% 
  mutate_if(is.character,
            stri_trans_general,
            id = "latin-ascii")

feature.cell.order.acute <- cell.data.hc.normalized %>% 
  filter(Time == "Acute") %>% 
  filter(HealthyInf == "Malaria infected") %>% 
  group_by(feature) %>% 
  summarise(order.median.acute = median(norm2hcpkl)) %>% 
  ungroup() %>% 
  arrange(desc(order.median.acute)) %>% 
  select(feature, order.median.acute)
```

##### Data prep plasma proteins

```{r}
## cohort data
Endemic_protein.data <- all.data %>% 
  mutate(Time = as.factor(gsub(".*\\|","",sample)),
         ID = gsub("\\|.*","",sample))  %>% 
  rename(Endemic = group) %>% 
  #select(ID,view,Time,value,feature) %>% 
  #inner_join(subjectTable,by="ID") %>% 
  select(ID,Endemic, Time, view, feature, value) %>% 
  mutate(unite = ifelse(view == "Olink","NPX","cells per uL blood (log2)"),
         Endemic = factor(Endemic, levels = c("primary_infected","previously_exposed")))

## calculate median of healthy ctrl
HC_protein.data <- protein.data.ctrl %>% 
  mutate(ID = gsub("_","",ID),
         view = Omics,
         Endemic = "Ctrl",
         Time = "Ctrl",
         unite = "NPX") %>% 
  dplyr::select(ID,Endemic,Time,view,feature,value,unite)

## join cohort and healthy control data
NEC <- rbind(Endemic_protein.data, HC_protein.data) %>%
  filter(view == "Olink") %>% 
  mutate(Endemic = as.factor(Endemic),
         Time = as.factor(Time)) %>% 
  mutate(Endemic = factor(Endemic, levels = c("primary_infected","previously_exposed","Ctrl")))

## join ctrl medians with patient data
protein.data.hc.normalized <- NEC %>% 
  full_join(NEC %>% 
              filter(Time == "Ctrl") %>%
              group_by(feature) %>% 
              summarise(hcNPX = median(value)) %>% 
              ungroup(),
            by = "feature") %>% 
  mutate(HealthyInf = as.factor(ifelse(Endemic == "Ctrl","Healthy","Malaria infected"))) %>% 
  mutate(norm2hcNPX = value - hcNPX) 

## order proteins based on median acute value in cohort
feature.protein.order.acute <- protein.data.hc.normalized %>% 
  filter(Time == "Acute") %>% 
  filter(HealthyInf == "Malaria infected") %>% 
  group_by(feature) %>% 
  summarise(order.median.acute = median(norm2hcNPX)) %>% 
  ungroup() %>% 
  arrange(desc(order.median.acute)) %>% 
  select(feature, order.median.acute) %>% 
  rownames_to_column("acute_order")


## calculate median of healthy ctrl
HC_cell.data <- facs.pkl.data.ctrl %>% 
  filter(gate != "id") %>% 
  mutate(ID = gsub("_","",sampleID),
         feature = gate.lab,
         view = "Facs",
         Endemic = "Ctrl",
         Time = "Ctrl",
         unite = "pkl") %>% 
  dplyr::select(ID,Endemic,Time,view,feature,value,unite) %>% 
  mutate(log.value = log(value),
         log2.value = log2(value +1),
         log10.value = log10(value)) %>% 
  select(ID,Endemic,Time,view,feature,log2.value) %>% 
  rename(value=log2.value) %>% 
  mutate_if(is.character,
            stri_trans_general,
            id = "latin-ascii")

```


```{r }
ifelse(isFALSE(dir.exists(paste0(result.dir,"/Cohort_vs_HealthyCtrl/"))), dir.create(paste0(result.dir,"/Cohort_vs_HealthyCtrl/"),recursive = TRUE),"HealthyCtrl Folder excists exists already!")
```

### Fig. S3A-D
* manuscript included only sample time points Acute and Y1

#### Immune cells

```{r }
for(TimePoint in c("Acute","D10","M1","M3","M6","Y1")){
  plot.data <- cell.data.hc.normalized %>%  
    filter(Time %in% c("Ctrl",TimePoint))
  
  plot.data.ordered <- feature.cell.order.acute %>%
    right_join(plot.data, by="feature")
  
  ## ======== Healthy vs Infected ====== ##
  stat.test <- plot.data.ordered %>%
    group_by(feature) %>%
    wilcox_test(norm2hcpkl ~ HealthyInf, p.adjust.method = "fdr") %>%
    adjust_pvalue() %>%
    add_significance("p.adj") %>% 
    inner_join(feature.cell.order.acute %>% select(feature,order.median.acute) %>% distinct(), by="feature") %>% 
    arrange(desc(order.median.acute)) %>% 
    add_xy_position(x="order.median.acute")
  
 # write.csv(stat.test %>% mutate(groups = as.character(groups)), paste0(result.dir,"Cohort_vs_HealthyCtrl/Facs_",TimePoint,"_HC_normalized.csv"))
  
  hc.norm <-
    plot.data.ordered %>% 
    ggplot(aes(x= fct_reorder(feature, order.median.acute), y= norm2hcpkl)) + 
    geom_hline(yintercept= 0, color = "grey20") +
    geom_boxplot(aes(fill=HealthyInf),
                 width = 1, 
                 position = position_dodge(1),
                 outlier.size = 0.2) +
    scale_fill_manual(values = c("grey50","#ca5572")) +
    labs(y = "cells/1000 live cells [log2] (normalized to healthy median)",
         x = "",
         title = TimePoint,
         caption = "Wilcoxon Tests with FDR correction") +
    theme_classic() +
    coord_flip() +
    theme_pubr() +
    theme(legend.position = "bottom") +
    stat_pvalue_manual(stat.test,
                       label = "p.adj.signif", 
                       y.position = min(plot.data.ordered$norm2hcpkl, na.rm = T)*1.05,
                       remove.bracket = T) 
  
  hc.norm# %>% ggsave(file = paste0(result.dir,"Cohort_vs_HealthyCtrl/Facs_",TimePoint,"_HC_normalized.pdf"), width = 6, height = 16, useDingbats = FALSE)
  
}
```

#### Plasma proteins

```{r}
for(TimePoint in c("Acute","D10","M1","M3","M6","Y1")){
  plot.data <- protein.data.hc.normalized %>%  
    filter(Time %in% c("Ctrl",TimePoint))
  
  plot.data.ordered <- feature.protein.order.acute %>%
    right_join(plot.data, by="feature")
  
  ## ======== Healthy vs Infected ====== ##
  stat.test <- plot.data.ordered %>%
    group_by(feature) %>%
    wilcox_test(norm2hcNPX ~ HealthyInf, p.adjust.method = "fdr") %>%
    adjust_pvalue() %>%
    add_significance("p.adj") %>% 
    inner_join(feature.protein.order.acute %>% select(feature,order.median.acute) %>% distinct(), by="feature") %>% 
    arrange(desc(order.median.acute)) %>% 
    add_xy_position(x="order.median.acute")
  
  #write.csv(stat.test %>% mutate(groups = as.character(groups)), paste0(result.dir,"Cohort_vs_HealthyCtrl/Olink_",TimePoint,"_HC_normalized.csv"))
  
  hc.norm <-
    plot.data.ordered %>% 
    ggplot(aes(x= fct_reorder(feature, order.median.acute), y= norm2hcNPX)) + 
    geom_hline(yintercept= 0, color = "grey20") +
    geom_boxplot(aes(fill=HealthyInf),
                 width = 1, 
                 position = position_dodge(1),
                 outlier.size = 0.2) +
    scale_fill_manual(values = c("grey50","#ca5572")) +
    labs(y = "NPX (normalized to healthy median)",
         x = "",
         title = TimePoint,
         caption = "Wilcoxon Tests with FDR correction") +
    theme_classic() +
    coord_flip() +
    theme_pubr() +
    theme(legend.position = "bottom") +
    stat_pvalue_manual(stat.test,
                       label = "p.adj.signif", 
                       y.position = min(plot.data.ordered$norm2hcNPX, na.rm = T)*1.05,
                       remove.bracket = T) 
  hc.norm #%>% ggsave(file = paste0(result.dir,"Cohort_vs_HealthyCtrl/Olink_",TimePoint,"_HC_normalized.pdf"), width = 6, height = 16, useDingbats = FALSE)
}
```


# == Clinical chemistry data ==

## Figure Supplementary 1

### Fig. S1C

```{r}
plot.input <- wbc.data %>% 
  mutate(Time=as.factor(Time)) %>% 
  select(sampleID,ID,Time,lpk,lymf,mono) %>% 
  gather(feature,value,lpk:mono) 

feature.labs <- c("Lymphocytes/1000 counts", "Lymphocytes", "Monocytes")
names(feature.labs) <- c("lpk", "lymf", "mono")

(pA <- plot.input %>% 
  ggplot(aes(x=Time, y=value, group=feature))+
  facet_wrap(~feature, ncol = 3, scales = "free_y",
             labeller = labeller(feature = feature.labs)) +
  geom_smooth(method="loess", color = "#6097ce") +
  labs(x = "Sample time point",
       y = "Cell count") +
   jtools::theme_nice() +
  theme_figpanel+
   theme(legend.position = "none",
         legend.key.size = unit(2, "cm"))
         )
```

### Fig. S1D
* cells/1000 live cell counts correlated with wbc-standardised cells per ul

```{r fig.asp=0.29, message=FALSE, warning=FALSE}
plot.input <- facs.pkl.data2 %>% 
  filter(value != "-Inf") %>% 
  mutate(rowname = paste(sampleID,"|",feature)) %>% 
  mutate(Time = as.factor(gsub(".*\\|","",sampleID))) %>% 
  select(rowname,value,Time) %>% rename("log2[cells per 1000 live]"=value) %>% 
  inner_join(facs.cpu.data %>% 
      mutate(rowname = paste(sampleID,"|",feature)) %>%
      select(rowname,value) %>% 
      filter(value != "-Inf") %>% 
      rename("log2[cells per ul]"=value),
    by="rowname") %>% 
  mutate(feature = gsub(".*\\|","",rowname),
         ID = gsub("\\|.*","",rowname))

(pB <- plot.input %>% 
    ggplot(aes(x=`log2[cells per 1000 live]`,y=`log2[cells per ul]`,color=Time)) +
    geom_point(alpha=0.5,size=0.5) +
    facet_wrap(~Time, ncol = 6) +
    stat_smooth(color="black",
                method = "lm",size=0.5,
                se = T,
                fullrange = T,
                alpha = 0.6) +
    stat_cor(aes(label = paste0("R = ", ..r..)),size=2,
             col="black",
             method = "spearman", 
             label.sep="\n", 
             output.type = "latex",
             hjust = 0,
             show.legend = F) +
    scale_color_manual(values=TIME_colors) +
    jtools::theme_nice() +
    theme_figpanel+
    theme(legend.position = "none",
          panel.spacing.x=unit(0.5, "lines"),
          legend.key.size = unit(2, "cm"))
  )

```

```{r}
#pdf("../plots/WBC_Corr_FigS1CD_revision_feb7.pdf", width = 6,height = 3, useDingbats = F,   title="main", paper = "a4", compress = T ) 
pA / pB 
#dev.off()#
```

