---
title: "Systems Immunology Malaria Traveller Cohort"
subtitle: "MEFISTO downstream analysis"
author: "Maximilian Julius Lautenbach"
date: '`r format(Sys.Date(), "%Y-%m-%d")`'
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: kate
    code_folding: hide
    df_print: paged
knit: (function(inputFile, encoding) {
          rmarkdown::render(inputFile,
                            encoding = encoding, 
                            output_file = paste0(
                              xfun::sans_ext(inputFile), '_', Sys.Date(), '.html'),
                                output_dir = "../results/reports/")})
---

```{r setup, include=FALSE}
result.dir <- paste0("results/figures/")
options(stringsAsFactors = FALSE) 
```

## required libraries
```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(viridis)
library(MOFA2)
library(RColorBrewer)
library(ComplexHeatmap)
library(circlize)
source("helper_fun.R")
```

# Preparation

### load trained mefisto model
* script for model training: ```MEFISTO.R```

```{r message=FALSE, warning=FALSE}
model <- load_model("../data/MEFISTO_model_updated.hdf5", load_interpol_Z = TRUE)
```


### load additional data 

```{r message=FALSE, warning=FALSE}
mefisto.input <- readRDS("../data/MEFISTO_input_data.Rds")
all.data <- readRDS("../data/ALL_input_data.Rds")
subjectTable <- read_csv("../data/TravellerCohort_SubjectTable.csv")
sampleTable <- read_csv("../data/TravellerCohort_SampleTable.csv")
```


### add meta data to model
* ```TravellerCohort_SubjectTable.csv```

```{r add sample meta data}
samples_metadata(model) <- samples_metadata(model) %>% 
  mutate(Time = as.factor(gsub(".*\\|","",sample)),
         ID = gsub("\\|.*","",sample)) %>% 
  inner_join(subjectTable, by=c("ID")) %>%
  distinct()
## rename view names
views_names(model) <- c("Immune cells", "Plasma proteins")
```

### extract model factors and weights

```{r}
factors.all <- get_factors(model,
                           factors = "all",
                           scale = FALSE,
                           as.data.frame = T) %>%
  inner_join(model@samples_metadata %>% select(-group), by="sample")
weights.all <- get_weights(model,
                           views = "all",
                           factors = "all", 
                           as.data.frame = TRUE)
```

# MEFISTO model downstream analysis

## Figure 2

### Fig. 2A - Variance explained

```{r}
(p1a <- as.data.frame(model@cache$variance_explained$r2_total) %>%
   rownames_to_column("View") %>% 
   gather(group,value,-View) %>% 
   mutate(group = factor(group, levels = c("primary_infected","previously_exposed"))) %>% 
   ggplot(aes(x=group, y= value, fill=View)) +
   geom_bar(stat="identity",position=position_dodge()) +
   scale_fill_manual(values = OMICS_colors) +
   labs(x = "",
        y = expression(paste("Total Var. ", r^{2}, " (%)")))+
   theme_classic(base_size = 6) +
   theme(legend.position = "top",
         axis.text.x = element_blank(),
         axis.ticks.x = element_blank()))
ggsave(file = paste0("../plots/total_variance_font12.pdf"),p1a, width = 5, height = 2.5 , useDingbats = FALSE)

(p1a2 <- plot_variance_explained(model) + 
    scale_fill_distiller(type = "seq",
                         palette = 1,
                         direction = 1,
                         values = NULL,
                         space = "Lab",
                         na.value = "grey50",
                         guide = "colourbar",
                         aesthetics = "fill") +
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_text(size = 6, angle = 45, hjust = 0.8)))
#legend.justification = "top"))
ggsave(file = paste0("../plots/variance_perfactor_font12.pdf"), p1a2, width = 7 ,height = 2, useDingbats = FALSE)


#pdf("../plots/Plot2a_upper.pdf", width = 2,height = 1,useDingbats = F,   title="main", paper = "a4", compress = T ) #colormodel = "cmyk"
p1a  + theme_figpanel + theme(legend.position = "none")
#dev.off()

#pdf("../plots/Plot2a_lower.pdf", width = 2,height = 2,useDingbats = F,   title="main", paper = "a4", compress = T ) #colormodel = "cmyk"
p1a2  + theme_figpanel #+ theme(legend.position = "right")
#dev.off()

```

### Fig. 2B - Visualisation of Factor1 over time after symptom onset

```{r}
factor1 <- get_factors(model, factors = c("Factor1"), scale = F, as.data.frame = T) %>% 
  inner_join(model@samples_metadata %>% select(-group), by="sample")

(p1b <- ggplot(factor1, 
               aes(x = weeks.po, y = value, color = Time)) + 
    geom_point(size=1) +
    geom_line(aes(group=ID),color="grey", alpha=0.15)+
    theme_classic(base_size = 6) +
    scale_color_manual(values=TIME_colors) +
    labs(x = "Time after onset (weeks)",
         y = "Factor 1 values",
         color = "Sample time point",
         fill = "weeks") +
    theme12 +
    theme(#legend.position='top',
      legend.title.align=0.5,
      legend.position =  c(0.6, 0.9)) +
    scale_x_continuous(breaks = pretty(factor1$weeks.po, n = 10)) +
    guides(colour=guide_legend(title="Sample Time Point",
                               override.aes=list(size=1),
                               nrow=2)))
 ggsave(paste0("../plots/Factor1_vs_weekspo.pdf"), p1b, width = 5, height = 5)
 
#pdf("../plots/Factor1_vs_weekspo_revision.pdf", width = 2,height = 2,useDingbats = F,   title="main", paper = "a4", compress = T ) #colormodel = "cmyk"
p1b + theme_figpanel
#dev.off()
```

### Fig. 2C - Lollipop-plots of Factor1 - driving features
* Top 10 positive and negative Factor 1 features

```{r}
## Immune cells
top10.feature.cells <- bind_rows(get_weights(model,
                                             views = "Immune cells",
                                             factors = 1,
                                             as.data.frame = TRUE,
                                             scale = T) %>%
                                   arrange(desc(value)) %>% 
                                   head(10) %>% mutate(direction = "positive"),
                                 get_weights(model,
                                             views = "Immune cells",
                                             factors = 1,
                                             as.data.frame = TRUE,
                                             scale = T) %>%
                                   arrange(value) %>% 
                                   head(10) %>% mutate(direction = "negative"))

## save plot data
write.csv(top10.feature.cells, "../data/Top10_Cells_F1.csv", row.names = F)

## lollipop plot 
(p1c1 <- top10.feature.cells %>% 
    mutate(direction = factor(as.factor(direction), levels=c("positive","negative"))) %>% 
    mutate(feature = as.character(feature)) %>% 
    ungroup() %>%  
    lollipop_plot() +
    labs(title = "Immune cells") +
    scale_y_continuous(limits = c(-1, 1), breaks = seq(-1, 1, by = 0.5)))
#ggsave(paste0("../plots/TopImmunecells_TopWeights_MOFAplot.pdf"),p1c1, height = 4.5, width = 5.2)

## Plasma proteins
top10.feature.proteins <- bind_rows(get_weights(model,
                                                views = "Plasma proteins",
                                                factors = 1,
                                                as.data.frame = TRUE,
                                                scale = T) %>%
                                      arrange(desc(value)) %>% 
                                      head(10) %>% mutate(direction = "positive"),
                                    get_weights(model,
                                                views = "Plasma proteins",
                                                factors = 1,
                                                as.data.frame = TRUE,
                                                scale = T) %>%
                                      arrange(value) %>% 
                                      head(10) %>% mutate(direction = "negative"))

## save plot data
write.csv(top10.feature.proteins, "../data/Top10_Proteins_F1.csv", row.names = F)

## lollipop plot 
(p1c2 <- top10.feature.proteins %>%
    mutate(direction = factor(as.factor(direction), levels=c("positive","negative"))) %>% 
    mutate(feature = as.character(feature)) %>% 
    ungroup() %>% 
    lollipop_plot() +
    labs(title = "Plasma proteins") +
    scale_y_continuous(limits = c(-0.5, 1), breaks = seq(-0.5, 1, by = 0.5)))
#ggsave(paste0("../plots/TopPlasmaproteins_TopWeights_MOFAplot.pdf"),p1c2, height = 4.5, width = 3)


library(patchwork)
#pdf("../plots/TopWeights_Lollipop.pdf", width = 3.5,height = 2,useDingbats = F,   title="Supplementary", paper = "a4", compress = T ) #colormodel = "cmyk"
p1c1 + theme_figpanel | p1c2 + theme_figpanel
#dev.off()

```

### Fig. 1D - Immune landscape heatmap 

#### Imputation of missing data points based on modified version of MOFA's heatmap function 
* modified function in ```helper_fun.R```

```{r message=FALSE, warning=FALSE}
data.prot.inputed <- plot_data_heatmap_ordered(object <- model,
                                               factor="Factor1",
                                               view="Plasma proteins",
                                               groups = "all",
                                               features = read_csv("../data/Top10_Proteins_F1.csv")  %>% pull(feature),
                                               imputed = TRUE,
                                               denoise = TRUE,
                                               max.value = NULL,
                                               min.value = NULL) %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("sample")

data.facs.inputed <- plot_data_heatmap_ordered(object <- model,
                                               factor="Factor1",
                                               view="Immune cells",
                                               groups = "all",
                                               features = read_csv("../data/Top10_Cells_F1.csv")  %>% pull(feature),
                                               imputed = TRUE,
                                               denoise = TRUE,
                                               max.value = NULL,
                                               min.value = NULL) %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("sample")

## save all data
data.all.inputed <- data.prot.inputed %>% 
  rename_with(~paste0("Protein_", .x), -c("sample")) %>% 
  inner_join(data.facs.inputed %>% 
               rename_with(~paste0("Cells_", .x), -c("sample")), 
             by="sample") %>% 
  right_join(samples_metadata(model) %>% select(sample,group,Time,days.po, weeks.po, weeks.po_warped), by="sample") %>% 
  select(sample,group,Time,days.po,weeks.po,weeks.po_warped,everything())
#write.csv(data.all.inputed,"../data/Factor1_inputed_samples.csv", row.names = F)
```

#### ComplexHeatmap on imputed data

```{r fig.asp=0.5, message=FALSE, warning=FALSE}
mat <- data.all.inputed %>% 
  arrange(weeks.po_warped) %>%
  select(-c("days.po","group","Time","weeks.po","weeks.po_warped")) %>%
  column_to_rownames("sample") %>% 
  t() 

## sample annotation
sample_anno <- data.all.inputed %>% #
  select(sample,group,Time, weeks.po_warped) %>% #days.po, weeks.po,
  column_to_rownames("sample") %>% 
  rename("Sample Time Point" = Time,
         "Weeks after onset warped" = weeks.po_warped,
         "Group" = group) 

## feature annotation
feature_anno <- data.frame("feature" = rownames(mat)) %>% 
  mutate(View = ifelse(grepl("Protein",feature),"Plasma proteins","Immune cells"))

rownames(feature_anno) <- rownames(feature_anno$feature)

feature_anno <- feature_anno %>% select(-feature)

## plotting heatmap
(hm <- mat %>% 
    na.omit() %>% 
    ComplexHeatmap::pheatmap(cluster_rows = T, 
                             cluster_cols = F,
                             show_colnames = F, 
                             show_rownames = T,
                             annotation_names_row = F,
                             annotation_colors = list("Sample Time Point" = TIME_colors,
                                                      "Weeks after onset warped"=viridis::viridis(80, direction = -1), 
                                                      "Group" = ENDEMIC_colors, "View" = OMICS_colors),
                             annotation_legend = T,
                             scale = "row",
                             annotation_row = feature_anno,
                             annotation_col = sample_anno,
                             clustering_distance_rows= "euclidean",
                             cellwidth = 3, 
                             cellheight = 6,
                             fontsize_row = 6,
                             treeheight_row =40,
                             name = "z-score",
                             fontsize = 12,
                             color = colorRamp2(c(-1,0,1), c("#4575b4","#ffffbf","#d73027"))))

#pdf(file=paste0("../plots/Heatmap_top10_imputed.pdf"), width=14.5, height = 10)
hm 
#dev.off()
```

## Figure Supplementary 2

### Fig. S2A sample coverage
```{r message=FALSE, warning=FALSE}
#pdf("../plots/SampleCoverage_Cohort_Overview.pdf", width = 6.5,height = 4,useDingbats = F,   title="Supplementary", paper = "a4", compress = T ) #colormodel = "cmyk"

plot_data_overview(model,
                   colors = OMICS_colors,
                   show_covariate = TRUE,
                   show_dimensions = TRUE) +
  theme(axis.text = element_text(size=6))
#dev.off()
#ggsave(paste0("../plots/SampleCoverage_Cohort_Overview.pdf"), width = 10, height = 6)
```

### Fig. S2B - Scatterplot Factor1 vs Factor2 

```{r fig.show="hold", message=FALSE, warning=FALSE, out.width="50%"}
(p1a3 <- factors.all %>% 
   spread(factor,value) %>% 
   ggplot(aes(x=Factor1,y=Factor2, col=days.po.log)) +
   geom_point() +
   scale_color_viridis_c(direction=-1) +
   labs(color="Days post onset\n    (log2 scale)") +
   
   theme_classic(base_size = 6) +
   theme(legend.direction = 'horizontal',
         legend.title.align=0.5,
         legend.position = "top") + #c(0.5, 0.9)) +
   theme_figpanel)
#pdf("../plots/Factor1vs2_dayspo.pdf", width = 2.5,height = 3,useDingbats = F,   title="Supplementary", paper = "a4", compress = T ) #colormodel = "cmyk"
p1a3
#dev.off()
#ggsave(paste0("../plots/Factor1vs2_dayspo.pdf"), width = 3.5, height = 4)
```

### Fig. S2C - Time point grouped Factor1 values over time after symptom onset

```{r}
#pdf("../plots/Factor1_facet_vs_weekspo.pdf", width = 6,height = 3,useDingbats = F,   title="Supplementary", paper = "a4", compress = T ) #colormodel = "cmyk"

ggplot() + 
  geom_point(data=select(factor1, -Time),aes(x = weeks.po, y = value), col="grey80") +
  geom_point(data=factor1,aes(x = weeks.po, y = value, color = Time)) +
  facet_grid(~Time) +
  geom_point(size=1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic(base_size = 6) +
  scale_color_manual(values=TIME_colors) +
  labs(x = "Time after onset (weeks)",
       y = "Factor 1 values",
       color = "Sample time point",
       fill = "weeks") +
  theme12 +
  theme(legend.position='bottom') +
  guides(colour=guide_legend(title="",
                             override.aes=list(size=2),
                             nrow=1))  +
    theme_figpanel 

#dev.off()
```


## Figure 4

### Fig. 4A

### 3.2.1 Smooth fit for groups

```{r fig.show="hold", message=FALSE, warning=FALSE, out.width="100%", fig.height=3, fig.width=3}
#pdf("../plots/Factor1_horizontal_smooth.pdf", width = 1.4,height = 1.4, useDingbats = F,   title="Supplementary", paper = "a4", compress = T ) #colormodel = "cmyk"

factor1 %>% 
  ggplot(aes(x=weeks.po,y=value,group=group)) +
  geom_point(aes(fill=group),size=1,shape=21, color="black") +
  geom_smooth(aes(color=group),se = F) +
  scale_color_manual(values = ENDEMIC_colors) +
  scale_fill_manual(values = ENDEMIC_colors) +
  labs(x = "Time after onset\n(weeks)",
       y = "Factor values",
       color = NULL,
       fill = NULL) +
  facet_wrap(~factor, scales = "free", ncol = 5) +
  theme_classic(base_size = 6) +
  update_geom_defaults("line",list(size=0.2)) +
  scale_x_continuous(trans = "pseudo_log",breaks = c(0,10,20,40,60)) + 
  theme_figpanel +
  theme(legend.position='none') 

#dev.off()

```




### Fig. 4C

```{r message=FALSE, warning=FALSE}
#pdf("../plots/gdTcellsF1_vs_Pseudotime.pdf", width = 1.4,height = 3,useDingbats = F,   title="Supplementary", paper = "a4", compress = T ) #colormodel = "cmyk"

plot_data_vs_cov(model, 
                 warped = T,
                 factor = 1,
                 features = c("Total gdT cells","Vd2- gdT cells","Vd2+ gdT cells"),
                 view = "Immune cells",
                 dot_size = 1) +
  facet_wrap(~feature, ncol = 1) +
  geom_smooth(aes(group = group, col = group),
              se=F) +
  theme_classic(base_size = 6) +
  scale_fill_manual(values = ENDEMIC_colors) +
  scale_color_manual(values = ENDEMIC_colors) +
   #scale_x_continuous(trans = "pseudo_log",breaks = c(0,10,20,40,60)) + 
  scale_x_continuous(breaks = c(0.5,52), labels = c("Acute","Convalescence")) +
  labs(x="",
       y="MEFISTO-scaled cell counts") +
  guides(fill = F,
         col = F) +
  update_geom_defaults("line",list(size=0.2)) +

  theme_figpanel +
  theme(legend.position='top',
        axis.ticks.x = element_blank()) 
#dev.off()
```


## Figure Supplementary 4

### Fig. S4A

```{r message=FALSE, warning=FALSE}
#pdf("../plots/Factor1_endemic_vs_timepoint.pdf", width = 6,height =3,useDingbats = F,   title="Supplementary", paper = "a4", compress = T ) #colormodel = "cmyk"
factor1 %>% 
  mutate(group = factor(as.factor(group), levels = c("primary_infected", "previously_exposed"))) %>% 
  ggplot(aes(x = group, y = value, color = group)) + 
  geom_jitter(alpha=0.5, width = 0.2) +
  geom_boxplot(alpha=0.5,show.legend = F) +
  facet_grid(~Time) +
  stat_compare_means(label = "p.format", show.legend = F,size=2, color="grey20") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic(base_size = 6) +
  scale_color_manual(values=ENDEMIC_colors, breaks = c("primary_infected", "previously_exposed")) +
  scale_fill_manual(values=ENDEMIC_colors, breaks = c("primary_infected", "previously_exposed")) +

  labs(x ="",
       y = "Factor 1 values",
       color = "",
       fill = "weeks",
       caption = "Wilcoxon Rank sum test") +
  jtools::theme_nice() +
  theme_figpanel+
  theme(legend.position = "right",
        panel.spacing.x=unit(0.5, "lines"),
       axis.text.x = element_blank(),
        axis.ticks = element_blank()) 
#dev.off()
```
